---
title: "PMI - mice "
author: "Cadiz, Todd, Olney"
date: "04/05/2022"
output:
  html_document:
    theme: cerulean
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: inline
---

# Set working directory
```{r working_directory}
knitr::opts_knit$set(root.dir = ".")
```

# Load libraries
```{r libraries, message=FALSE, warnings=FALSE}
# load libraries
library(cowplot)  # plot_grid()
library(DoubletFinder)  # paramSweep()
library(dplyr)  # ungroup()
library(ggrepel) # geom_text_repel()
library(ggplot2)  # ggplot()
library(grid)  # grid.arrange()
library(gridExtra)  # grid.arrange()
library(harmony) # RunHarmony()
library(reshape2) # melt()
library(Seurat)  # Read10X_h5()
library(Seurat.utils) # RenameGenesSeurat()
library(stringr) # str_match()
```

# Set variables and thresholds
```{r set_variables}
pathToRef <- "/research/labs/neurology/fryer/projects/references/mouse/refdata-gex-mm10-2020-A"
tissue <- "Brain"
replicate <- "1"
treatment <- "PMI"
# genotype
WT <- "WT"
PS19 <- "PS19"
control <- "0_hr"
treatment_3hr <- "3_hr"
treatment_6hr <- "6_hr"
treatment_color <- "blue"
control_color <- "gray"
WT_treatment_color_3hr <- "#6699FF" # light blue grey
WT_treatment_color_6hr <- "#0033FF" # dark blue
PS19_treatment_color_3hr <- "#CC99FF" # light blue grey
PS19_treatment_color_6hr <- "#9933FF" # dark blue
sample_colors <- c(control_color,WT_treatment_color_3hr, WT_treatment_color_6hr, control_color, PS19_treatment_color_3hr, PS19_treatment_color_6hr)
treatment_colors <- c(control_color, treatment_color)
myContrasts <- c("6 hour - 0 hour")
tool <- "cellranger"
```

```{r set_thresholds}
nCount.min <- 250
nFeature.min <- 250
complexity.cutoff <- 0.85
mt.cutoff <- 25
hb.cutoff <- 5
```

# Save functions
These functions with help simultaneously save plots as a png and pdf.
```{r save_functions}
saveToPDF <- function(...) {
    d = dev.copy(pdf,...)
    dev.off(d)
}
saveToPNG <- function(...) {
    d = dev.copy(png,...)
    dev.off(d)
}
```

# Setup seurat object
Using CellBender filtered output.
```{r seurat_object}
prefix <- "../../cellranger/scRNA/"
suffix <- "/outs/filtered_feature_bc_matrix.h5"

if (tissue == "Brain" && file.exists("../../rObjects/cell_replicate_1_merged_h5.rds")) {
  PMI <- readRDS("../../rObjects/cell_replicate_1_merged_h5.rds")
} else if (tissue == "Brain") {
  # individual sample objects
  WT_0_hr <- CreateSeuratObject(Read10X_h5(paste0(prefix,"WT_0_hr",suffix)))
  WT_3_hr <- CreateSeuratObject(Read10X_h5(paste0(prefix,"WT_3_hr",suffix)))
  WT_6_hr <- CreateSeuratObject(Read10X_h5(paste0(prefix,"WT_6_hr",suffix)))
  PS19_0_hr <- CreateSeuratObject(Read10X_h5(paste0(prefix,"PS19_0_hr",suffix)))
  PS19_3_hr <- CreateSeuratObject(Read10X_h5(paste0(prefix,"PS19_3_hr",suffix)))
  PS19_6_hr <- CreateSeuratObject(Read10X_h5(paste0(prefix,"PS19_6_hr",suffix)))
 
  # merge objects
  PMI <- merge(x = WT_0_hr, 
                y = c(WT_3_hr, WT_6_hr, PS19_0_hr, PS19_3_hr, PS19_6_hr), 
                add.cell.ids = c("WT_0_hr", "WT_3_hr", "WT_6_hr", "PS19_0_hr", "PS19_3_hr", "PS19_6_hr"), 
                project = paste0("PMI replicate group ", replicate, " - Single Cell"))
  
  # cleanup and save
  remove(WT_0_hr, WT_3_hr, WT_6_hr, PS19_0_hr, PS19_3_hr, PS19_6_hr)
  saveRDS(PMI, "../../rObjects/cell_replicate_1_merged_h5.rds")
  
} 

# preview
PMI
```

nCount_RNA = total number of transcripts (UMIs) in a single cell 
nFeature_RNA = number of unique genes (features)

```{r change_idents}
# create sample column
barcodes <- colnames(PMI)
pattern <- "(.+)_[ACGT]+-(\\d+)"
sample <- str_match(barcodes, pattern)[,2]
table(sample)
PMI$sample <- factor(sample, levels = c("WT_0_hr", "WT_3_hr", "WT_6_hr", "PS19_0_hr", "PS19_3_hr", "PS19_6_hr"))
table(PMI$sample)  # check
Idents(PMI) <- PMI$sample
```

Add treatment column to metadata
```{r treatment_column}
# create treatment column
treat <- gsub("WT_0_hr", control, PMI$sample)
treat <- gsub("WT_3_hr", treatment_3hr, treat)
treat <- gsub("WT_6_hr", treatment_6hr, treat)
treat <- gsub("PS19_0_hr", control, treat)
treat <- gsub("PS19_3_hr", treatment_3hr, treat)
treat <- gsub("PS19_6_hr", treatment_6hr, treat)

PMI$treatment <- factor(treat, levels = c("0_hr","3_hr", "6_hr"))
table(PMI$treatment)
```
Add genotype column to metadata
```{r genotype_column}
# create treatment column
geno <- gsub("WT_0_hr", WT, PMI$sample)
geno <- gsub("WT_3_hr", WT, geno)
geno <- gsub("WT_6_hr", WT, geno)
geno <- gsub("PS19_0_hr", PS19, geno)
geno <- gsub("PS19_3_hr", PS19, geno)
geno <- gsub("PS19_6_hr", PS19, geno)

PMI$genotype <- factor(geno, levels = c("WT","PS19"))
table(PMI$genotype)
```


```{r QC_columns}
# cell.complexity
PMI$cell.complexity <- log10(PMI$nFeature_RNA) / log10(PMI$nCount_RNA)

# percent MT
grep("^mt-",rownames(PMI@assays$RNA@counts),value = TRUE)
mt.genes <- c("mt-Nd1", "mt-Nd2", "mt-Co1", "mt-Co2", "mt-Atp8", "mt-Atp6", "mt-Co3", "mt-Nd3",
              "mt-Nd4l", "mt-Nd4", "mt-Nd5", "mt-Nd6", "mt-Cytb")
PMI$percent.mt <- PercentageFeatureSet(PMI, pattern = "^mt-")

# ribosomal proteins begin with 'RPL' or 'RSL' for this annotation file
PMI$percent.ribo <- PercentageFeatureSet(PMI, pattern = "^R[sp]l")

# percent.hb - hemoglobin proteins begin with 'HB' or 'HBP' 
PMI$percent.hb <- PercentageFeatureSet(PMI, pattern = "^Hb[^(p)]")
```

# Pre-filtering QC
## Number of cells
```{r prefiltering_cells_per_sample}
# Visualize the number of cell counts per sample
data <- as.data.frame(table(PMI$sample))
colnames(data) <- c("sample","frequency")

ncells1 <- ggplot(data, aes(x = sample, y = frequency, fill = sample)) + 
  geom_col() +
  theme_classic() +
  geom_text(aes(label = frequency), 
            position=position_dodge(width=0.9), 
            vjust=-0.25) +
  scale_fill_manual(values = sample_colors) + 
  scale_y_continuous(breaks = seq(0,55000, by = 5000), limits = c(0,55000)) +
  ggtitle("Raw: cells per sample") +
  theme(legend.position =  "none") + 
  theme(axis.text.x = element_text(angle = 45, hjust=1))
ncells1
```

## Density plots
```{r prefiltering_density}
# set graphical parameter
par(mfrow = c(3,1))

# Visualize nCount_RNA
den1 <- ggplot(PMI@meta.data,
       aes(color = sample,
           x = nCount_RNA,
           fill = sample)) +
  geom_density(alpha = 0.2) +
  theme_classic() +
  scale_x_log10() +
  scale_color_manual(values = sample_colors) +
  scale_fill_manual(values = sample_colors) +
  xlab("nCount_RNA") +
  ylab("Density") +
  geom_vline(xintercept = nCount.min)

# Visualize percent.mt
den2 <- ggplot(PMI@meta.data,
       aes(color = sample,
           x = percent.mt,
           fill = sample)) +
  geom_density(alpha = 0.2) +
  theme_classic() +
  scale_x_continuous(n.breaks = 4) +
  geom_vline(xintercept = mt.cutoff) +
  scale_color_manual(values = sample_colors) +
  scale_fill_manual(values = sample_colors) +
  xlab("% Mitochondrial Genes") +
  ylab("Density")

# Visualize cell complexity
# Quality cells are usually above 0.85
den3 <- ggplot(PMI@meta.data,
       aes(color = sample,
           x = cell.complexity,
           fill = sample)) +
  geom_density(alpha = 0.2) +
  theme_classic() +
  scale_color_manual(values = sample_colors) +
  scale_fill_manual(values = sample_colors) +
  xlab("Cell Complexity (log10(nFeature/nCount))") +
  ylab("Density") +
  geom_vline(xintercept = complexity.cutoff)

# Arrange graphs in grid
plots1 <- list(den1,den2,den3)
layout1 <- rbind(c(1),c(2),c(3))
grid1 <- grid.arrange(grobs = plots1, layout_matrix = layout1)
```

## Violin plots
```{r prefiltering_violins}
# nFeature, nCount, and cell.complexity violins
v1 <- VlnPlot(PMI,
              features = c("nFeature_RNA", "nCount_RNA","cell.complexity"),
              ncol = 3,
              group.by = 'sample',
              cols = sample_colors,
              pt.size = 0)
v1

#  percent violins
v2 <- VlnPlot(PMI,
              features = c("percent.mt","percent.ribo","percent.hb"),
              ncol = 3,
              group.by = 'sample',
              cols = sample_colors,
              pt.size = 0)
v2
```

```{r, echo=FALSE,eval=FALSE}
# save v1
v1
path <- paste0("../../results/vlnPlots/",treatment,"_",tolower(tissue),"_nFeature_nCount_complexity_raw")
saveToPDF(paste0(path, ".pdf"), width = 6, height = 4)
saveToPNG(paste0(path, ".png"), width = 6, height = 4, unit = "in", res = 300)
dev.off()

# save v2
v2
path <- paste0("../../results/vlnPlots/",treatment,"_",tolower(tissue),"_percent_raw")
saveToPDF(paste0(path, ".pdf"), width = 6, height = 4)
saveToPNG(paste0(path, ".png"), width = 6, height = 4, unit = "in", res = 300)
dev.off()

# cleanup
remove(v1,v2)
```

## Scatter plots
```{r prefiltering_scatter1}
s1 <- ggplot(
  PMI@meta.data,
  aes(x = nCount_RNA, y = nFeature_RNA, color = percent.mt)) + 
  geom_point() + 
  stat_smooth(method=lm) +
	scale_x_log10() +   	
  scale_y_log10() + 
  theme_classic() +
  geom_vline(xintercept = nCount.min) + 
  geom_hline(yintercept = nFeature.min) + 
  facet_wrap(~sample) +
  scale_colour_gradient(low = "gray90", high = "black", limits =c(0,100))
  #geom_rect(aes(xmin=300, xmax=300, ymin=1000,
   #             ymax=3000), color="transparent", fill="orange", alpha=0.3)
s1
```

```{r echo=FALSE,eval=FALSE}
# save
s1
path <- paste0("../../results/scatterPlots/",treatment,"_",tolower(tissue),"_nFeature_vs_nCount_raw")
saveToPDF(paste0(path, ".pdf"), width = 4, height = 4)
saveToPNG(paste0(path, ".png"), width = 4, height = 4, unit = "in", res = 300)
dev.off()

# cleanup
remove(s1)
```

```{r prefiltering_scatter2}
s2 <- FeatureScatter(PMI,
               feature1 = "nCount_RNA",
               feature2 = "percent.mt",
               group.by = 'sample',
               cols = sample_colors,
               shuffle = TRUE)
s2
```

```{r,echo=FALSE,eval=FALSE}
# save
s2
path <- paste0("../../results/scatterPlots/",treatment,"_",tolower(tissue),"_nCount_vs_percentMT_raw")
saveToPDF(paste0(path, ".pdf"), width = 4, height = 4)
saveToPNG(paste0(path, ".png"), width = 4, height = 4, unit = "in", res = 300)
dev.off()

# cleanup
remove(s2)
```

# Filtering
## Cell-level filtering
We want to be careful filtering because removing things can easily lead to
misinterpretation.  For example, cells with high percent.mt could actually
just be involved in respiratory processes. \

We will filter based on 6 conditions: \
-- nCount_RNA > 500
-- nFeature_RNA > 250
-- cell.complexity > 0.85
-- percent.mt < 1
-- percent.hb < 5

And removing MT genes as to not alter down stream differential expression
```{r cell_filtering}
# filter
PMI.filtered <- subset(PMI,
                        subset = (nCount_RNA > nCount.min) &
                          (nFeature_RNA > nFeature.min) &
                          (cell.complexity > complexity.cutoff) &
                          (percent.mt < mt.cutoff))

# print cells removed
print(paste0(dim(PMI)[2] - dim(PMI.filtered)[2]," cells removed"))
```

## Gene-level filtering
Remove lowly expressed genes.  We will keep genes that have at least 1 count in
10 cells.
```{r gene_filtering, warning=FALSE, message=FALSE}
# filter genes
counts <- GetAssayData(object = PMI.filtered, slot = "counts")
nonzero <- counts > 0  # produces logical
keep <- Matrix::rowSums(nonzero) >= 10  # sum the true/false
counts.filtered <- counts[keep,]  # keep certain genes

# overwrite PMI.filtered
PMI.filtered <- CreateSeuratObject(counts.filtered, 
                                    meta.data = PMI.filtered@meta.data)

# print features removed
print(paste0(dim(counts)[1] - dim(counts.filtered)[1], " features removed"))
```


## Doublet filtering
https://github.com/chris-mcginnis-ucsf/DoubletFinder\

Heterotypic - doublets derived from transcriptionally distinct cells. 
DoubletFinder works best on this type of doublet. \

Homotopic - Transcriptionally similar cell doublets. DoubletFinder does not work 
as great on this type of doublet.\

pANN - proportion of artificial nearest neighbors (pANN)\

BCMVN - mean-variance normalized bimodality coefficient of pANN distributions 
produced during pN -pK parameter sweeps. The BCMVN may be used to identify the 
pK parameter.\

Overview of steps:\
A. Prepare each sample\
B. pK Identification (no ground-truth) - defines the PC neighborhood size used 
   to compute pANN\
C. Homotypic Doublet Proportion Estimate - homotypic doublets may not be a 
   problem depending on the type of analysis you are performing. If you have 
   some doublets of the same type and their counts are normalized, they will 
   generally represent the profile of single cells of the same type.\
D. DoubletFinder\
E. Visualize where the doublets are located\

```{r split_object}
# split object by sample
PMI.split <- SplitObject(PMI.filtered, split.by = "sample") 
```

```{r doubletFinder, message=FALSE, warning=FALSE}
for (i in 1:length(PMI.split)) {
  # normalize and find PCs
  print(i)
  PMI_sample <- NormalizeData(PMI.split[[i]])
  sampleID <- levels(droplevels(PMI_sample@meta.data$sample))
  PMI_sample <- FindVariableFeatures(PMI_sample, selection.method = "vst", nfeatures = 2000)
  PMI_sample <- ScaleData(PMI_sample)
  PMI_sample <- RunPCA(PMI_sample)
  
  # get significant PCs
  stdv <- PMI_sample[["pca"]]@stdev
  sum.stdv <- sum(PMI_sample[["pca"]]@stdev)
  percent.stdv <- (stdv / sum.stdv) * 100
  cumulative <- cumsum(percent.stdv)
  co1 <- which(cumulative > 90 & percent.stdv < 5)[1]
  co2 <- sort(which((percent.stdv[1:length(percent.stdv) - 1] - 
                       percent.stdv[2:length(percent.stdv)]) > 0.1), 
              decreasing = T)[1] + 1
  min.pc <- min(co1, co2)
  min.pc
  
  # run umap
  PMI_sample <- RunUMAP(PMI_sample, dims = 1:min.pc, reduction = "pca")
  
  # cluster
  PMI_sample <- FindNeighbors(object = PMI_sample, dims = 1:min.pc)                           
  PMI_sample <- FindClusters(object = PMI_sample, resolution = 0.2)
  
  # Assign identity of clusters
  Idents(object = PMI_sample) <- "RNA_snn_res.0.4"
  d1 <- DimPlot(PMI_sample,
        reduction = "umap",
        label = TRUE,
        label.size = 6)
  path <- paste0("../../results/doubletFinder/",treatment,"_",tolower(tissue),
               "_doubletFinder_UMAP_res0.02_",sampleID)
  pdf(paste0(path, ".pdf"), width = 5, height = 4)
  print(d1)
  dev.off()
  
  # number of cells in each cluster
  n_cells <- FetchData(PMI_sample, vars = c("ident")) %>% dplyr::count(ident) %>%tidyr::spread(ident, n)
  
  ## pK Identification (no ground-truth) 
  sweep.res.list <- paramSweep_v3(PMI_sample, PCs = 1:min.pc, sct = FALSE)
  sweep.stats <- summarizeSweep(sweep.res.list, GT = FALSE)
  bcmvn <- find.pK(sweep.stats)
  
  # Optimal pK for any scRNA-seq data can be manually discerned as maxima in BCmvn distributions
  bcmvn_max <- bcmvn[which.max(bcmvn$BCmetric),]
  pK_value <- bcmvn_max$pK
  pK_value <- as.numeric(levels(pK_value))[pK_value]
  
  # Homotypic Doublet Proportion Estimate 
  annotations <- PMI_sample@meta.data$seurat_clusters
  homotypic.prop <- modelHomotypic(annotations) 
  nExp_poi <- round(pK_value*nrow(PMI_sample@meta.data))
  nExp_poi.adj <- round(nExp_poi*(1-homotypic.prop))
  # Run DoubletFinder with varying classification  
  PMI_sample <- doubletFinder_v3(PMI_sample, PCs = 1:min.pc, 
                pN = 0.25, pK = pK_value, nExp = nExp_poi.adj, 
                reuse.pANN = FALSE, sct = FALSE)
  
  # set DF class for calling doublets
  DF_class <- PMI_sample@meta.data[, grep("DF.classifications",colnames(PMI_sample@meta.data)),]
  DF_class[which(DF_class == "Doublet")] <- "Doublet"
  table(DF_class)
  
  # table showing the number of doublets and singlets
  write.table(table(DF_class), paste0("../../results/doubletFinder/",treatment,"_",tolower(tissue),
               "_doubletFinder_table_",sampleID), sep = "\t", 
              row.names = FALSE, quote = FALSE)
  PMI_sample@meta.data[,"CellTypes_DF"] <- DF_class
  
  # plot
  d2 <- DimPlot(PMI_sample, group.by="CellTypes_DF", reduction="umap",
          order=c("Coll.Duct.TC","Doublet"), 
          cols=c("#66C2A5","black"))
  path <- paste0("../../results/doubletFinder/",treatment,"_",tolower(tissue),
               "_doubletFinder_UMAP_",sampleID)
  pdf(paste0(path, ".pdf"), width = 5,height = 4)
  print(d2)
  dev.off()
  
  # plot
  f1 <- FeaturePlot(PMI_sample, 
            reduction = "umap", 
            features = c("nFeature_RNA", "nCount_RNA", 
                         "cell.complexity", "percent.mt"),
            pt.size = 0.4, 
            order = TRUE,
            label = TRUE)
  path <- paste0("../../results/doubletFinder/",treatment,"_",tolower(tissue),
               "_doubletFinder_FeaturePlot_",sampleID)
  pdf(paste0(path, ".pdf"), width = 7, height = 7)
  print(f1)
  dev.off()
  
  #only keep singlets
  PMI_sample_singlets <- subset(PMI_sample, subset = CellTypes_DF == "Singlet")
  
  # inspect
  d3 <- DimPlot(PMI_sample_singlets, group.by="CellTypes_DF", reduction="umap",
          order=c("Coll.Duct.TC","Doublet"),
          cols=c("#66C2A5","black"))
  path <- paste0("../../results/doubletFinder/",treatment,"_",tolower(tissue),
               "_doubletFinder_UMAP_singlets_",sampleID)
  pdf(paste0(path, ".pdf"), width = 5, height = 4)
  print(d3)
  dev.off()
  
  # number of cells in each cluster per and post removing doublets
  n_cells_singlets <- FetchData(PMI_sample_singlets, vars = c("ident")) %>% dplyr::count(ident) %>% tidyr::spread(ident, n)
  n_cells_singlets
  ncells_per_cluster <- rbind(n_cells, n_cells_singlets)
  row.names(ncells_per_cluster) <- c("Doublets and singlets", "Singlets only")
  ncells_per_cluster
  difference <- diff(as.matrix(ncells_per_cluster))
  difference <- as.data.frame(difference)
  row.names(difference) <- c("difference")
  cbind(difference, ncells_per_cluster)
  write.table(ncells_per_cluster, paste0(
    "../../results/doubletFinder/",treatment,"_",tolower(tissue),
    "_doubletFinder_table_ncells_per_cluster",sampleID, ".txt"), sep = "\t", 
    row.names = FALSE, quote = FALSE)
  
  # plot the number of cells in each cluster per and post doubletFinder
  ncell_matrix <- as.matrix(ncells_per_cluster)
  ncells_melt <- melt(ncell_matrix)
  colnames(ncells_melt) <- c("doublet type","cluster","number of cells")
  ncell_max <- ncells_melt[which.max(ncells_melt$`number of cells`),]
  ncell_max_value <- ncell_max$`number of cells`
  cellmax <- ncell_max_value + 800 # so that the figure doesn't cut off the text
  b1 <- ggplot(ncells_melt, aes(x = factor(cluster), y = `number of cells`,
                          fill = `doublet type`)) + 
    geom_bar(stat="identity", colour="black", width=1, position = position_dodge(width=0.8)) +
    geom_text(aes(label = `number of cells`), 
              position=position_dodge(width=0.9), vjust=-0.25, angle = 45, hjust=-.01) + 
    theme_classic() + scale_fill_manual(values = c("gray", "#66C2A5")) +
    ggtitle("Number of cells per cluster") +  xlab("cluster") +
    theme(axis.text.x = element_text(angle = 45, hjust=1)) +
    scale_y_continuous(limits = c(0,cellmax))
  path <- paste0("../../results/doubletFinder/",treatment,"_",tolower(tissue),
               "_doubletFinder_barplot_ncells_per_cluster",sampleID)
  pdf(paste0(path, ".pdf"), width = 7,height = 5)
  print(b1)
  dev.off()
  f2 <- FeaturePlot(PMI_sample_singlets, 
            reduction = "umap", 
            features = c("nFeature_RNA", "nCount_RNA", 
                         "cell.complexity", "percent.mt"),
            pt.size = 0.4, 
            order = TRUE,
            label = TRUE)
  path <- paste0("../../results/doubletFinder/",treatment,"_",tolower(tissue),
               "_doubletFinder_FeaturePlot_singlets",sampleID)
  pdf(paste0(path, ".pdf"), width = 7,height = 7)
  print(f2)
  dev.off()
  
  # put the PMI together again
  PMI.split[[i]] <- PMI_sample_singlets
}
```

```{r}
# converge PMI.split
PMI.singlets <- merge(x = PMI.split[[1]],
                       y = c(PMI.split[[2]], PMI.split[[3]],PMI.split[[4]], PMI.split[[5]], PMI.split[[6]]),
                       project = paste0("LPS PMI ", tissue, " Single Cell"))

# print how many cells removed
print(paste0(dim(PMI.filtered)[2] - dim(PMI.singlets)[2]," cells removed"))

# how many removed if we had an upper nCount and nFeature
PMI.upper <- subset(PMI.singlets,
                     subset = (nCount_RNA < 10000) & (nFeature_RNA < 5000))
print(paste0(dim(PMI.filtered)[2] - dim(PMI.upper)[2],
             " cells would have been removed if upper bound applied"))

# overwrite PMI.filtered
PMI.filtered <- PMI.singlets

# reset levels
PMI.filtered$treatment <- factor(PMI.filtered$treatment,
                                  levels = c("0_hr","3_hr", "6_hr"))
PMI.filtered$sample <- factor(PMI.filtered$sample,
                               levels = c("WT_0_hr", "WT_3_hr", "WT_6_hr", "PS19_0_hr", "PS19_3_hr", "PS19_6_hr"))

# cleanup
remove(PMI.singlets, PMI.upper, PMI.split, PMI_sample_singlets, PMI_sample)
remove(n_cells,n_cells_singlets,ncell_matrix,ncell_max,ncells_per_cluster,ncells_melt)
remove(sweep.res.list, sweep.stats,bcmvn,bcmvn_max,difference)
remove(d1,d2,d3,f1,f2)
remove(counts,counts.filtered, nonzero)
```

## Mitochondrial gene filtering
```{r}
# remove mt.genes
counts <- GetAssayData(object = PMI.filtered, slot = "counts")
keep <- !rownames(counts) %in% mt.genes # false when mt.gene
counts.filtered <- counts[keep,]

# overwrite PMI.filtered
PMI.filtered <- CreateSeuratObject(counts.filtered, 
                                    meta.data = PMI.filtered@meta.data)

# print features removed
print(paste0(dim(counts)[1] - dim(counts.filtered)[1], " features removed"))
```

```{r,echo=FALSE}
saveRDS(PMI.filtered, paste0("../../rObjects/",tolower(tissue),"_filtered.rds"))
PMI.filtered <- readRDS(paste0("../../rObjects/",tolower(tissue),"_filtered.rds"))
Idents(PMI.filtered) <- PMI.filtered$sample
```

## Histogram
```{r}
# User params
goi <- "Xist"
threshold <- 0.1

# Subset data
log2.threshold <- log2(threshold + 0.01)
counts.df <- FetchData(PMI.filtered, vars = goi)
colnames(counts.df) <- "counts"
log2.counts.df <- log2(counts.df + 0.01)

# Histogram
title <- paste0("LPS Brain Nuclei: ", goi, "\nnCount_RNA > ", threshold)
hist1 <- ggplot(counts.df, aes(x = counts)) + 
  geom_histogram(bins = 100, fill = "gray", color = "black") + 
  labs(title = title, x=NULL, y=NULL) +
  xlab(paste0(goi, " nCount_RNA")) + ylab("# of Samples") + theme_bw() +
  geom_vline(xintercept = threshold, col = "blue") +
  annotate("rect", 
           xmin = -Inf,
           xmax = threshold, 
           ymin = 0, 
           ymax=Inf, 
           alpha=0.2, 
           fill="chocolate4") +
  annotate("rect", 
           xmin = threshold,
           xmax = Inf, 
           ymin = 0, 
           ymax=Inf, 
           alpha=0.2, 
           fill="deepskyblue")

# Histogram log transformed
hist2 <- ggplot(log2.counts.df, aes(x = counts)) + 
  geom_histogram(bins = 100, fill = "gray", color = "black") + 
  labs(title = title, x=NULL, y=NULL) +
  xlab(paste0("Log2(",goi, " nCount_RNA)")) + ylab("# of Samples") + theme_bw() +
  geom_vline(xintercept = log2.threshold, col = "blue") +
  annotate("rect", 
           xmin = -Inf,
           xmax = log2.threshold, 
           ymin = 0, 
           ymax=Inf, 
           alpha=0.2, 
           fill="chocolate4") +
  annotate("rect", 
           xmin = log2.threshold,
           xmax = Inf, 
           ymin = 0, 
           ymax=Inf, 
           alpha=0.2, 
           fill="deepskyblue")

# plot
plots1 <- list(hist1,hist2)
layout1 <- rbind(c(1),c(2))
grid1 <- grid.arrange(grobs = plots1, layout_matrix = layout1)

# number removed
table(counts.df$counts > threshold)
```
## Xist
remove cells that do not express Xist
```{r}
# which cells have a count for Xist?
cellsXist <- WhichCells(PMI.filtered, slot = 'counts', expression = Xist > 0 )

# percent xist in each cell 
PMI.filtered$percent.xist <- PercentageFeatureSet(PMI.filtered, features = "Xist")

vXist <- VlnPlot(PMI.filtered,
              features = c("percent.xist"),
              ncol = 3,
              group.by = 'sample',
              cols = sample_colors,
              pt.size = 0)
vXist

# remove cells less than 0 percent Xist expresison
xist.filtered <- subset(PMI.filtered,
                        subset = percent.xist > 0)

# print cells removed
print(paste0(dim(PMI.filtered)[2] - dim(xist.filtered)[2]," cells removed"))
# print features removed
print(paste0(dim(PMI.filtered)[1] - dim(xist.filtered)[1], " features removed"))

remove(PMI.filtered)
# overwrite PMI.filtered
PMI.filtered <- xist.filtered
```

```{r,echo=FALSE,eval=FALSE}
# save
grid1 <- grid.arrange(grobs = plots1, layout_matrix = layout1)
path <- paste0("../../results/histogram/",treatment,"_",tolower(tissue), "_",
               goi,"_histogram")
saveToPDF(paste0(path, ".pdf"), width = 6, height = 4)
saveToPNG(paste0(path, ".png"), width = 6, height = 4, unit = "in", res = 300)
dev.off()

# cleanup
remove(log2.threshold,counts.df,log2.counts.df, hist1,hist2,plots1,layout1,grid1)
```

## Number of cells
```{r number_cells2}
# Visualize the number of cell counts per sample
data <- as.data.frame(table(PMI.filtered$sample))
colnames(data) <- c("sample","frequency")

ncells2 <- ggplot(data, aes(x = sample, y = frequency, fill = sample)) + 
  geom_col() +
  theme_classic() +
  geom_text(aes(label = frequency), 
            position=position_dodge(width=0.9), 
            vjust=-0.25) +
  scale_fill_manual(values = sample_colors) + 
  scale_y_continuous(breaks = seq(0,55000, by = 5000), limits = c(0,55000)) +
  ggtitle("Filtered: cells per sample") +
  theme(legend.position =  "none") + 
  theme(axis.text.x = element_text(angle = 45, hjust=1))

# Arrange graphs in grid
plots2 <- list(ncells1,ncells2)
layout2 <- cbind(c(1),c(2))
grid2 <- grid.arrange(grobs = plots2, layout_matrix = layout2)
```

```{r,echo=FALSE,eval=FALSE}
# save
grid2 <- grid.arrange(grobs = plots2, layout_matrix = layout2)
path <- paste0("../../results/nCells/",treatment,"_",tolower(tissue), 
               "_cells_per_sample")
saveToPDF(paste0(path, ".pdf"), width = 6, height = 4)
saveToPNG(paste0(path, ".png"), width = 6, height = 4, unit = "in", res = 300)
dev.off()

# cleanup
remove(ncells1,ncells2,plots2,layout2,grid2)
```
## Density plots
```{r postfiltering_density}
# set graphical parameter
par(mfrow = c(3,1))

# Visualize the number of counts per cell
den4 <- ggplot(PMI.filtered@meta.data,
       aes(color = sample,
           x = nCount_RNA,
           fill = sample)) +
  geom_density(alpha = 0.2) +
  theme_classic() +
  scale_x_log10() +
  scale_color_manual(values = sample_colors) +
  scale_fill_manual(values = sample_colors) +
  xlab("nCount_RNA") +
  ylab("Density") +
  geom_vline(xintercept = nCount.min)

# Visualize percent.mt
den5 <- ggplot(PMI.filtered@meta.data,
       aes(color = sample,
           x = percent.mt,
           fill = sample)) +
  geom_density(alpha = 0.2) +
  theme_classic() +
  scale_x_log10() +
  scale_color_manual(values = sample_colors) +
  scale_fill_manual(values = sample_colors) +
  xlab("% Mitochondrial Genes") +
  ylab("Density") +
  geom_vline(xintercept = mt.cutoff)

# Visualize cell complexity
# Quality cells are usually above 0.80
den6 <- ggplot(PMI.filtered@meta.data,
       aes(color = sample,
           x = cell.complexity,
           fill = sample)) +
  geom_density(alpha = 0.2) +
  theme_classic() +
  scale_x_log10() +
  scale_color_manual(values = sample_colors) +
  scale_fill_manual(values = sample_colors) +
  xlab("Cell Complexity (log10(nFeature/nCount))") +
  ylab("Density") +
  geom_vline(xintercept = complexity.cutoff)

# Arrange graphs in grid
plots3 <- list(den1,den2,den3,den4,den5,den6)
layout3 <- rbind(c(1,4),c(2,5),c(3,6))
grid3 <- grid.arrange(grobs = plots3, layout_matrix = layout3)
```

```{r, echo=FALSE,eval=FALSE}
# save
grid3 <- grid.arrange(grobs = plots3, layout_matrix = layout3)
path <- paste0("../../results/density/",treatment,"_",tolower(tissue), 
               "_density")
saveToPDF(paste0(path, ".pdf"), width = 6, height = 4)
saveToPNG(paste0(path, ".png"), width = 6, height = 4, unit = "in", res = 300)
dev.off()

# cleanup
remove(den1,den2,den3,den4,den5,den6,plots3,layout3,grid3)
```

## Violin plots
```{r postfiltering_violins}
# nFeature, nCount, and cell.complexity violins
v3 <- VlnPlot(PMI.filtered,
              features = c("nFeature_RNA", "nCount_RNA","cell.complexity"),
              ncol = 3,
              group.by = 'sample',
              cols = sample_colors,
              pt.size = 0)
v3

#  percent violins
v4 <- VlnPlot(PMI.filtered,
              features = c("percent.mt","percent.ribo","percent.hb"),
              ncol = 3,
              group.by = 'sample',
              cols = sample_colors,
              pt.size = 0)
v4

#  percent violins
v5 <- VlnPlot(PMI.filtered,
              features = c("percent.xist"),
              group.by = 'sample',
              cols = sample_colors,
              pt.size = 0)
v5
```

```{r, echo=FALSE,eval=FALSE}
# save
v3
path <- paste0("../../results/vlnPlots/",treatment,"_",tolower(tissue), 
               "_nFeature_nCount_complexity_filtered")
saveToPDF(paste0(path, ".pdf"), width = 6, height = 4)
saveToPNG(paste0(path, ".png"), width = 6, height = 4, unit = "in", res = 300)
dev.off()

# save
v4
path <- paste0("../../results/vlnPlots/",treatment,"_",tolower(tissue), 
               "_percent_filtered")
saveToPDF(paste0(path, ".pdf"), width = 6, height = 4)
saveToPNG(paste0(path, ".png"), width = 6, height = 4, unit = "in", res = 300)
dev.off()

v5
path <- paste0("../../results/vlnPlots/",treatment,"_",tolower(tissue), 
               "_percent_xist_filtered")
saveToPDF(paste0(path, ".pdf"), width = 6, height = 4)
saveToPNG(paste0(path, ".png"), width = 6, height = 4, unit = "in", res = 300)
dev.off()
# cleanup
remove(v3,v4,v5)
```

## Scatter plots
```{r postfiltering_scatter1}
s3 <- ggplot(
  PMI.filtered@meta.data,
  aes(x = nCount_RNA, y = nFeature_RNA, color = percent.mt)) + 
  geom_point() + 
  stat_smooth(method=lm) +
	scale_x_log10() +   	
  scale_y_log10() + 
  theme_classic() +
  geom_vline(xintercept = nCount.min) + 
  geom_hline(yintercept = nFeature.min) + 
  facet_wrap(~sample) +
  scale_colour_gradient(low = "gray90", high = "black", limits =c(0,100))
  #geom_rect(aes(xmin=300, xmax=300, ymin=1000,
   #             ymax=3000), color="transparent", fill="orange", alpha=0.3)
s3
```

```{r,echo=FALSE,eval=FALSE}
# save
s3
path <- paste0("../../results/scatterPlots/",treatment,"_",tolower(tissue),
               "_nFeature_vs_nCount_filtered")
saveToPDF(paste0(path, ".pdf"), width = 4, height = 4)
saveToPNG(paste0(path, ".png"), width = 4, height = 4, unit = "in", res = 300)
dev.off()

# cleanup
remove(s3)
```

```{r postfiltering_scatter2}
s4 <- FeatureScatter(PMI.filtered,
               feature1 = "nCount_RNA",
               feature2 = "percent.mt",
               group.by = 'sample',
               cols = sample_colors,
               shuffle = TRUE)
s4
```

```{r,echo=FALSE,eval=FALSE}
# save
s4
path <- paste0("../../results/scatterPlots/",treatment,"_",tolower(tissue),
               "_nCount_vs_percentMT_filtered")
saveToPDF(paste0(path, ".pdf"), width = 4, height = 4)
saveToPNG(paste0(path, ".png"), width = 4, height = 4, unit = "in", res = 300)
dev.off()

# cleanup
remove(s4)
```

## Box plot
```{r boxplot}
# Visualize the distribution of genes detected per cell via boxplot
b1 <- ggplot(PMI.filtered@meta.data,
       aes(x = sample, 
           y = log10(nFeature_RNA), 
           fill=sample)) + 
  geom_boxplot() + 
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  theme(plot.title = element_text(hjust = 0.5, face="bold")) +
  ggtitle("Unique Genes / Cell / Sample") +
  scale_color_manual(values = sample_colors) +
  scale_fill_manual(values = sample_colors) +
  xlab("Sample")
b1
```

```{r,echo=FALSE,eval=FALSE}
# save
b1
path <- paste0("../../results/boxPlots/",treatment,"_",tolower(tissue),
               "_nFeature_per_sample")
saveToPDF(paste0(path, ".pdf"), width = 4, height = 4)
saveToPNG(paste0(path, ".png"), width = 4, height = 4, unit = "in", res = 300)
dev.off()

# cleanup
remove(b1)
```

# Top transcripts
```{r top_transcripts}
df <- data.frame(row.names = rownames(PMI.filtered))
df$rsum <- rowSums(x = PMI.filtered, slot = "counts")
df$gene_name <- rownames(df)
df <- df[order(df$rsum,decreasing = TRUE),]
head(df, 10)
```

```{r,echo=FALSE,eval=FALSE}
write.table(df, paste0("../../results/topTranscripts/", tolower(tissue), "_abundant_transcripts.txt"),
            quote = FALSE,
            row.names = FALSE)
```


# Next steps

For something to be informative, it needs to exhibit variation, but not all 
variation is informative. The goal of our clustering analysis is to keep the 
major sources of variation in our dataset that should define our cell types, 
while restricting the variation due to uninteresting sources of variation 
(sequencing depth, cell cycle differences, mitochondrial expression, batch 
effects, etc.). Then, to determine the cell types present, we will perform a 
clustering analysis using the most variable genes to define the major sources of 
variation in the dataset.

1. Explain sources of unwanted variation: cell cycle, percent.mt (cell stress)
2. Normalizing and regressing out sources of unwanted variation: sctransform
3. Integration (OPTIONAL STEP): recommended when comparing groups, uses shared
   sources of high variation to identify shared sub-populations across conditions,
   we visualize after integration to ensure it's good before clustering
   - Stuart and Butler et al. (2018)
4. Clustering cells: cluster cells based on similarity of their gene expression
   profiles, cells are assigned to a cluster based on their PCA scores based on
   expression of integrated highly variable genes
5. Cluster quality evaluation: see what clusters aren't influenced by sources
   of uninteresting variation, check major PCs are driving clusters, explore
   cell identities by looking a known markers

# Cell cycle

The most common biological data correction is to remove the effects of the cell 
cycle on the transcriptome. This data correction can be performed by a simple 
linear regression against a cell cycle score.

Check cell cycle phase BEFORE doing sctransform.  Counts to need to be 
comparable between cells and each cell has a different number for nCount_RNA.  

```{r summary}
# summary of counts per cell
summary(PMI.filtered@meta.data$nCount_RNA)
```

Use the NormalizeData() function with the argument LogNormalze to account for 
sequencing depth. nCount_RNA for each gene is divided by the total nCount_RNA 
for that cell.  This is done for all cells.  This number is then multiplied by 
the scale.factor so we don't have to deal with a tiny number. This number is 
then natural-log transformed using log1p. log1p is the natural logarithm (base 
e) of 1 + count. The 1 will prevent taking the log of 0.


```{r log_normalize}
PMI.phase <- NormalizeData(PMI.filtered,
                            scale.factor = 10000, # default
                            normalization.method = "LogNormalize" # default
                            )
```

Give each cell a score based on expression of G1, G2/M, and S phase markers.  A 
list of markers is provided for humans.  We will need to convert human gene names to mouse gene names.
We will use the CellCycleScoring()
function in seurat.

Below is a resource for acquiring cell markers for other organisms 
https://hbctraining.github.io/scRNA-seq_online/lessons/cell_cycle_scoring.html

G1 ~10 hrs
S  ~5-6 hrs
G2 ~3-4 hrs
M  ~2 hrs

G1 (10 hrs) > G2/M (5-6 hrs) = S (5-6 hrs)

If the score is negative for both S.Score and G2M.Score the phase is G1.
Otherwise the the greatest positive value between S.Score and G2M.Score
determines the phase.

Using a human list for mouse 
https://github.com/satijalab/seurat/issues/2493 
```{r fx_convert_gene_names}
# Basic function to convert human to mouse gene names
convertHumanGeneList <- function(x){
require("biomaRt")
human = useMart("ensembl", dataset = "hsapiens_gene_ensembl")
mouse = useMart("ensembl", dataset = "mmusculus_gene_ensembl")
genesV2 = getLDS(attributes = c("hgnc_symbol"), filters = "hgnc_symbol", values = x , mart = human, attributesL = c("mgi_symbol"), martL = mouse, uniqueRows=T)
humanx <- unique(genesV2[, 2])
# Print the first 6 genes found to the screen
print(head(humanx))
return(humanx)
}
#genes <- convertHumanGeneList(humGenes)
```

```{r phase}
m.s.genes <- convertHumanGeneList(cc.genes.updated.2019$s.genes)
m.g2m.genes <- convertHumanGeneList(cc.genes.updated.2019$g2m.genes)


# score cells for cell cycle
PMI.phase <- CellCycleScoring(PMI.phase,
                               g2m.features = m.g2m.genes,
                               s.features = m.s.genes,
                               set.ident = TRUE)
```

## Bar graph
```{r}
cellcyclecount_barplot <- 
  as_tibble(PMI.phase[[]]) %>%
  ggplot(aes(Phase, fill = Phase)) + geom_bar()
cellcyclecount_barplot
```

```{r,echo=FALSE,eval=FALSE}
# save
cellcyclecount_barplot
path <- paste0("../../results/cellCycle/",treatment,"_",tolower(tissue),
               "_cellcycle_bargraph")
saveToPDF(paste0(path, ".pdf"), width = 4, height = 4)
saveToPNG(paste0(path, ".png"), width = 4, height = 4, unit = "in", res = 300)
dev.off()

# cleanup
remove(cellcyclecount_barplot)
```

## Pie chart
```{r}
# pie point 
cellcyclecount_piepoint <- 
  as_tibble(PMI.phase[[]]) %>%
  ggplot(aes(x=S.Score, y=G2M.Score, color=Phase)) + 
  geom_point()
cellcyclecount_piepoint
```

```{r,echo=FALSE,eval=FALSE,message=FALSE,warning=FALSE}
# save
cellcyclecount_piepoint
path <- paste0("../../results/cellCycle/",treatment,"_",tolower(tissue),
               "_cellcycle_piechart")
saveToPDF(paste0(path, ".pdf"), width = 4, height = 4)
saveToPNG(paste0(path, ".png"), width = 4, height = 4, unit = "in", res = 300)
dev.off()

# cleanup
remove(cellcyclecount_piepoint)
```

## Top variable genes
```{r scale_data}
# Identify the most variable genes
PMI.phase <- FindVariableFeatures(PMI.phase,
                                   selection.method = "vst",  # default vst
                                   nfeatures = 2000,  # default 2000
                                   verbose = FALSE)

# view top variable genes
top40 <- head(VariableFeatures(PMI.phase), 40)
top40
```

```{r, warning=FALSE, message=FALSE}
# plot variable features with labels
VarFeatPlot <- VariableFeaturePlot(PMI.phase, cols = c("gray47","red"))
VarFeatPlotLabel <- LabelPoints(plot = VarFeatPlot, 
                    points = top40, repel = TRUE, fontface="italic", 
                    xnudge = 0, ynudge = 0, max.overlaps = 12)
VarFeatPlotLabel
```
## Percent top variable genes
```{r percent_ttr}
PMI.filtered$percent.Xist <- PercentageFeatureSet(PMI.filtered, feature = "Xist")
PMI.filtered$percent.Ttr <- PercentageFeatureSet(PMI.filtered, feature = "Ttr")
PMI.filtered$percent.Apoe <- PercentageFeatureSet(PMI.filtered, feature = "Apoe")

# nFeature, nCount, and cell.complexity violins
v1 <- VlnPlot(PMI.filtered,
              features = c("percent.Xist", "percent.Ttr", "percent.Apoe"),
              ncol = 3,
              group.by = 'sample',
              cols = sample_colors,
              pt.size = 0)
v1

```

```{r,echo=FALSE,eval=FALSE,message=FALSE,warning=FALSE}
# save
path <- paste0("../../results/cellCycle/",treatment,"_",tolower(tissue),
               "_cellcycle_variance_vs_expression")
VarFeatPlotLabel
saveToPDF(paste0(path, ".pdf"), width = 4, height = 4)
saveToPNG(paste0(path, ".png"), width = 4, height = 4, unit = "in", res = 300)
dev.off()

# cleanup
remove(VarFeatPlot,VarFeatPlotLabel)
```

## Mean-variance
```{r,message=FALSE,warning=FALSE}
# The variability information can be accessed using the HVFInfo method. 
# The names of the variable features can be accessed with VariableFeatures().
variance.data <- as_tibble(HVFInfo(PMI.phase),rownames = "Gene")
variance.data <- variance.data %>% mutate(hypervariable=Gene %in% VariableFeatures(PMI.phase))

# We can plot out a graph of the variance vs mean and highlight the selected genes 
# this way, we can see whether we think we’re likely to capture what we need.
subset_data <- subset(variance.data, hypervariable == TRUE)
varGeneslog <- variance.data %>% 
  ggplot(aes(log(mean),log(variance),color=hypervariable)) + 
  geom_point() + 
  scale_color_manual(values=c("black","red")) +  geom_text_repel(
    data = subset_data, max.overlaps = 20,
    aes(
      x = log(mean),
      y = log(variance),
      label = Gene, 
      fontface="italic",),segment.alpha = 1,size = 4) +
  theme(legend.position="bottom")
varGeneslog
```

```{r,echo=FALSE,eval=FALSE,message=FALSE,warning=FALSE}
# save
varGeneslog
path <- paste0("../../results/cellCycle/",treatment,"_",tolower(tissue),
               "_cellcycle_variance_vs_mean")
saveToPDF(paste0(path, ".pdf"), width = 4, height = 4)
saveToPNG(paste0(path, ".png"), width = 4, height = 4, unit = "in", res = 300)
dev.off()

# cleanup
remove(varGeneslog,variance.data)
```

## PCA
See if the cell cycle is a major source of variation using PCA. Choose the most 
variable gene features (we have already done), then e the data. We scale the 
data because highly expressed genes exhibit the highest amount of variation and 
we don’t want our ‘highly variable genes’ only to reflect high expression.

vst: First, fits a line to the relationship of log(variance) and log(mean) using 
local polynomial regression (loess). Then, feature values are standardized using 
the observed mean and expected variance (given by the fitted line). Feature 
variance is then calculated on the standardized values after clipping to a 
maximum (see clip.max parameter).

The ScaleData() function in Seurat will adjust gene expressions so that the mean 
expression in each cell is 0.  It will also scale each gene to give a variance 
of 1 for each cell.

```{r}
# Scale the counts
PMI.phase <- ScaleData(PMI.phase)
PMI.phase@assays
PMI.phase.pca <- RunPCA(PMI.phase, features = c(m.s.genes, m.g2m.genes))
DimPlot(PMI.phase.pca)
```

If the plots for each phase look very similar to each other, do not regress out
variation due to cell cycle.  You can plot PC1 vs PC2 before and after 
regression to see how effective it was. G1 (10 hrs) > G2/M (5-6 hrs) = S (5-6 hrs)

```{r view_phase, message=FALSE}
# Perform PCA
PMI.phase <- RunPCA(PMI.phase)

# Plot the PCA colored by cell cycle phase
cycle.pca <- DimPlot(PMI.phase,
        reduction = "pca",
        group.by= "Phase",
        split.by = "Phase")
cycle.pca
```

```{r,echo=FALSE,eval=FALSE,message=FALSE,warning=FALSE}
# save
cycle.pca
path <- paste0("../../results/cellCycle/",treatment,"_",tolower(tissue),
               "_cellcycle_PCA")
saveToPDF(paste0(path, ".pdf"), width = 4, height = 4)
saveToPNG(paste0(path, ".png"), width = 4, height = 4, unit = "in", res = 300)
dev.off()

# cleanup
remove(cycle.pca)
```

# SCTransform
Now, we can use the SCTransform method as a more accurate method of normalizing, 
estimating the variance of the raw filtered data, and identifying the most 
variable genes. Variation in sequencing depth (total nCount_RNA per cell) is 
normalized using a regularized negative binomial model.  ??Variance is also 
adjusted based on pooling information across genes with similar abundances??

Sctransform automatically accounts for cellular sequencing depth by regressing 
out sequencing depth (nUMIs). However, if there are other sources of 
uninteresting variation identified in the data during the exploration steps we 
can also include these. We observed little to no effect due to cell cycle phase 
and so we chose not to regress this out of our data. We observed some effect of 
mitochondrial expression and so we choose to regress this out from the data.

Since we have four samples in our dataset (from two conditions), we want to keep 
them as separate objects and transform them as that is what is required for 
integration. We will first split the cells in seurat.phase object by sample. 

```{r split_object2}
# Split seurat object by timepoint to perform SCT on all samples
PMI.split <- SplitObject(PMI.phase, split.by = "sample")
```

Now we will use a ‘for loop’ to run the SCTransform() on each sample, and 
regress out mitochondrial expression by specifying in the vars.to.regress 
argument of the SCTransform() function.

Before we run this for loop, we know that the output can generate large R 
objects/variables in terms of memory. If we have a large dataset, then we might 
need to adjust the limit for allowable object sizes within R (Default is 500 * 
1024 ^ 2 = 500 Mb) using the following code:

```{r sct, message=FALSE, warning=FALSE}
options(future.globals.maxSize = 4000 * 1024^5)

for (i in 1:length(PMI.split)) {
  print(paste0("Sample ", i))
  PMI.split[[i]] <- SCTransform(PMI.split[[i]]
                                #,
                                # vars.to.regress = c("percent.mt") 
                                 )
}
```

NOTE: By default, after normalizing, adjusting the variance, and regressing out 
uninteresting sources of variation, SCTransform will rank the genes by residual 
variance and output the 3000 most variant genes. If the dataset has larger cell 
numbers, then it may be beneficial to adjust this parameter higher using the 
variable.features.n argument.

Note, the last line of output specifies “Set default assay to SCT”. We can view 
the different assays that we have stored in our seurat object.

A thread about whether or not regress out batch:
https://github.com/satijalab/seurat/issues/3270
It is suggested to not regress out batch, and instead use a data integration method 

```{r check_split}
# Check
PMI.split
```

# Integration
## Run harmony
Condition-specific clustering of cells indicates that we need to integrate 
the cells across conditions to ensure that cells of the same cell type cluster 
together. \

To integrate, use the shared highly variable genes from each condition 
identified using SCTransform. Then, integrate conditions to overlay cells that 
are similar or have a “common set of biological features” between groups. \

Now, using our SCTransform object as input, let’s perform the integration across 
conditions. \

First, we need to specify that we want to use all of the 3000 most variable 
genes identified by SCTransform for the integration. By default, this function 
selects the top 2000 genes. \

```{r, warning=FALSE, message=FALSE}
# Choose the features to use when integrating multiple datasets. 
# will use nfeatures as 3000 as defined by running SCTransform above
var.features <- SelectIntegrationFeatures(object.list = PMI.split, 
                                          nfeatures = 3000)

# merge the PMI
PMI.sct.merged <- merge(x = PMI.split[[1]],
                         y = c(PMI.split[[2]],PMI.split[[3]],PMI.split[[4]],PMI.split[[5]], PMI.split[[6]]),
                         project = paste0("LPS PMI ", tissue, " Single Cell"))

# define the variable features 
VariableFeatures(PMI.sct.merged) <- var.features

# run PCA on the merged object
PMI.sct.merged <- RunPCA(object = PMI.sct.merged, assay = "SCT")

# run harmony to harmonize over samples 
PMI.integrated <- RunHarmony(object = PMI.sct.merged, 
                              group.by.vars = "sample", 
                              assay.use = "SCT",
                              reduction = "pca", 
                              plot_convergence = TRUE)
```

## Check output
```{r check_harmony}
# first put the PMI back in place
Idents(PMI.integrated) <- PMI.integrated$sample
PMI.integrated$treatment <- factor(PMI.integrated$treatment,
                                    levels = c("0_hr","3_hr", "6_hr"))
PMI.integrated$sample <- factor(PMI.integrated$sample,
                                 levels = c("WT_0_hr", "WT_3_hr", "WT_6_hr", "PS19_0_hr", "PS19_3_hr", "PS19_6_hr"))
# check the embedding
harmony_embeddings <- Embeddings(PMI.integrated, 'harmony')
harmony_embeddings[1:5, 1:5]

# check the PCA plot
p1 <- DimPlot(object = PMI.integrated, 
              reduction = "harmony",
              group.by = "treatment", 
              cols = treatment_colors) + NoLegend()
p2 <- VlnPlot(object = PMI.integrated, 
              features = "harmony_1", 
              group.by = "treatment", 
              pt.size = 0, 
              cols = treatment_colors) + NoLegend()
plot_grid(p1,p2)
```

## Top variable features
Top 20 variable features
```{r variable_genes}
top20 <- PMI.integrated@assays$SCT@var.features[1:20]
top20
```

## PCA plot
After integration, to visualize the integrated data we can use dimensionality 
reduction techniques, such as PCA and Uniform Manifold Approximation and 
Projection (UMAP). While PCA will determine all PCs, we can only plot two at a 
time. In contrast, UMAP will take the information from any number of top PCs to 
arrange the cells in this multidimensional space. It will take those distances 
in multidimensional space, and try to plot them in two dimensions. In this way, 
the distances between cells represent similarity in expression.

To generate these visualizations with the harmony output, use reduction = "harmony"

```{r plot_PCA}
# Plot PCA
pca1 <- DimPlot(PMI.integrated,
        reduction = "harmony",
        split.by = "treatment",
        group.by = "treatment",
        cols = treatment_colors)
pca1

pca2 <- DimPlot(PMI.integrated,
        reduction = "harmony",
        split.by = "sample",
        group.by = "sample",
        cols = sample_colors)
pca2

pca3 <- DimPlot(PMI.integrated,
        reduction = "harmony",
        group.by = "sample",
        shuffle = TRUE)
pca3
```

```{r,echo=FALSE,eval=FALSE,message=FALSE,warning=FALSE}
# save pca1
path <- paste0("../../results/PCA/",treatment,"_",tolower(tissue),
               "_PCA_split_group_color_group")
pca1
saveToPDF(paste0(path, ".pdf"), width = 4, height = 4)
saveToPNG(paste0(path, ".png"), width = 4, height = 4, unit = "in", res = 300)
dev.off()

# save pca2
path <- paste0("../../results/PCA/",treatment,"_",tolower(tissue),
               "_PCA_split_sample_color_group")
pca2
saveToPDF(paste0(path, ".pdf"), width = 4, height = 4)
saveToPNG(paste0(path, ".png"), width = 4, height = 4, unit = "in", res = 300)
dev.off()

# save pca3
path <- paste0("../../results/PCA/",treatment,"_",tolower(tissue),
               "_PCA_split_sample_color_unique")
pca3
saveToPDF(paste0(path, ".pdf"), width = 4, height = 4)
saveToPNG(paste0(path, ".png"), width = 4, height = 4, unit = "in", res = 300)
dev.off()

# cleanup
remove(pca1,pca2,pca3)
```

# Find significant PCs

To overcome the extensive technical noise in the expression of any single gene 
for scRNA-seq data, Seurat assigns cells to clusters based on their PCA scores 
derived from the expression of the integrated most variable genes, with each PC 
essentially representing a “metagene” that combines information across a 
correlated gene set. Determining how many PCs to include in the clustering step 
is therefore important to ensure that we are capturing the majority of the 
variation, or cell types, present in our dataset.

```{r view_PCs}
# Printing out the most variable genes driving PCs
print(x = PMI.integrated[["pca"]], 
      dims = 1:10, 
      nfeatures = 10)
```

Quantitative approach to an elbow plot
- The point where the principal components only contribute 5% of standard 
  deviation and the principal components cumulatively contribute 90% of the 
  standard deviation.
- The point where the percent change in variation between the consecutive PCs is 
  less than 0.1%.

First metric
```{r metric1}
# Determine percent of variation associated with each PC
stdv <- PMI.integrated[["pca"]]@stdev
sum.stdv <- sum(PMI.integrated[["pca"]]@stdev)
percent.stdv <- (stdv / sum.stdv) * 100

# Calculate cumulative percents for each PC
cumulative <- cumsum(percent.stdv)

# Determine which PC exhibits cumulative percent greater than 90% and
# and % variation associated with the PC as less than 5
co1 <- which(cumulative > 90 & percent.stdv < 5)[1]
co1
```

Second metric
```{r metric2}
# Determine the difference between variation of PC and subsequent PC
co2 <- sort(which(
  (percent.stdv[1:length(percent.stdv) - 1] - 
     percent.stdv[2:length(percent.stdv)]) > 0.1), 
  decreasing = T)[1] + 1

# last point where change of % of variation is more than 0.1%.
co2
```

Choose the minimum of these two metrics as the PCs covering 
the majority of the variation in the data.
```{r minimum_PCs}
# Minimum of the two calculation
min.pc <- min(co1, co2)
min.pc
```

## Elbow plot
Use min.pc we just calculated to generate the clusters. We can plot the elbow 
plot again and overlay the information determined using our metrics:
```{r quantitative_elbow}
# Create a dataframe with values
plot_df <- data.frame(pct = percent.stdv, 
           cumu = cumulative, 
           rank = 1:length(percent.stdv))

# Elbow plot to visualize 
  ggplot(plot_df, aes(cumulative, percent.stdv, label = rank, color = rank > min.pc)) + 
  geom_text() + 
  geom_vline(xintercept = 90, color = "grey") + 
  geom_hline(yintercept = min(percent.stdv[percent.stdv > 5]), color = "grey") +
  theme_bw()
```

# UMAP
```{r UMAP, message=FALSE, warning=FALSE}
# Run UMAP
PMI.integrated <- RunUMAP(PMI.integrated,
                           dims = 1:min.pc,
                           reduction = "harmony",
                           n.components = 3) # set to 3 to use with VR
```

```{r plot_UMAP}
# plot UMAP and color based on treatment
DimPlot(PMI.integrated,
        group.by = "treatment",
        split.by = "treatment",
        shuffle = TRUE,
        cols = treatment_colors)
```

# Cluster the cells

Seurat uses a graph-based clustering approach, which embeds cells in a graph 
structure, using a K-nearest neighbor (KNN) graph (by default), with edges drawn 
between cells with similar gene expression patterns. Then, it attempts to 
partition this graph into highly interconnected ‘quasi-cliques’ or ‘communities’ 
[Seurat - Guided Clustering Tutorial].

We will use the FindClusters() function to perform the graph-based clustering. 
The resolution is an important argument that sets the “granularity” of the 
downstream clustering and will need to be optimized for every individual experiment. 
For datasets of 3,000 - 5,000 cells, the resolution set between 0.4-1.4 generally 
yields good clustering. Increased resolution values lead to a greater number of 
clusters, which is often required for larger datasets.

The FindClusters() function allows us to enter a series of resolutions and will 
calculate the “granularity” of the clustering. This is very helpful for testing 
which resolution works for moving forward without having to run the function for 
each resolution.

```{r, echo=FALSE, eval=TRUE}
# save preclustering object
saveRDS(PMI.integrated, paste0("../../rObjects/",tolower(tissue),"_preclustering.rds"))
#PMI.integrated <- readRDS(paste0("../../rObjects/",tolower(tissue),"_preclustering.rds"#))
```


```{r find_neighbors, message=FALSE, warning=FALSE}
# Determine the K-nearest neighbor graph
PMI.unannotated <- FindNeighbors(object = PMI.integrated, 
                                 assay = "SCT", # set as default after SCTransform
                                 reduction = "harmony",
                                 dims = 1:min.pc)

# Determine the clusters for various resolutions
PMI.unannotated <- FindClusters(object = PMI.unannotated,
                                 algorithm = 1, # 1= Louvain
                                 resolution = seq(0.1,0.8,by=0.1))
```

```{r save_object_integrated,echo=FALSE,eval=TRUE}
# Save integrated seurat object
saveRDS(PMI.unannotated, paste0("../../rObjects/",tolower(tissue),"_unannotated.rds"))
#PMI.unannotated <- readRDS(paste0("../../rObjects/",tolower(tissue),"_unannotated.rds"))
```

```{r}
colors <- c("dodgerblue","yellow","firebrick1","gold","gray40","lightgray",
            "cyan","chocolate4","pink","orange","darkgreen","purple","lightgray","tan")
DimPlot(PMI.unannotated,
        group.by = "SCT_snn_res.0.1",
        label = TRUE,
        cols = colors)
```


# Explore resolutions
```{r}
# 0.1
umap0.1 <- DimPlot(PMI.unannotated,
        group.by = "SCT_snn_res.0.1",
        label = TRUE)
umap0.1

# 0.2
umap0.2 <- DimPlot(PMI.unannotated,
        group.by = "SCT_snn_res.0.2",
        label = TRUE)
umap0.2

# 0.3
umap0.3 <- DimPlot(PMI.unannotated,
        group.by = "SCT_snn_res.0.3",
        label = TRUE)
umap0.3

# 0.4
umap0.4 <- DimPlot(PMI.unannotated,
        group.by = "SCT_snn_res.0.4",
        label = TRUE)
umap0.4
```

```{r,echo=FALSE,eval=FALSE,message=FALSE,warning=FALSE}
# save umap0.1
path <- paste0("../../results/UMAP/unannotated/",treatment,"_",tolower(tissue),
               "_UMAP_unannotated_res0.1")
umap0.1
saveToPDF(paste0(path, ".pdf"), width = 4, height = 4)
saveToPNG(paste0(path, ".png"), width = 4, height = 4, unit = "in", res = 300)
dev.off()

# save umap0.2
path <- paste0("../../results/UMAP/unannotated/",treatment,"_",tolower(tissue),
               "_UMAP_unannotated_res0.2")
umap0.2
saveToPDF(paste0(path, ".pdf"), width = 4, height = 4)
saveToPNG(paste0(path, ".png"), width = 4, height = 4, unit = "in", res = 300)
dev.off()

# save umap0.3
path <- paste0("../../results/UMAP/unannotated/",treatment,"_",tolower(tissue),
               "_UMAP_unannotated_res0.3")
umap0.3
saveToPDF(paste0(path, ".pdf"), width = 4, height = 4)
saveToPNG(paste0(path, ".png"), width = 4, height = 4, unit = "in", res = 300)
dev.off()

# save umap0.4
path <- paste0("../../results/UMAP/unannotated/",treatment,"_",tolower(tissue),
               "_UMAP_unannotated_res0.4")
umap0.4
saveToPDF(paste0(path, ".pdf"), width = 4, height = 4)
saveToPNG(paste0(path, ".png"), width = 4, height = 4, unit = "in", res = 300)
dev.off()
```

# Clustering QC
## Treatment, sample, phase
```{r umap_treatment}
# treatment
u1 <- DimPlot(PMI.unannotated, 
        label = FALSE, 
        group.by = "SCT_snn_res.0.1",
        split.by = "treatment") +
  NoLegend()
u1

# sample
u2 <- DimPlot(PMI.unannotated, 
        label = FALSE,
        group.by = "SCT_snn_res.0.1",
        split.by = "sample") +
  NoLegend()
u2

# phase
u3 <- DimPlot(PMI.unannotated, 
        label = FALSE,
        group.by = "SCT_snn_res.0.1",
        split.by = "Phase") +
  NoLegend()
u3
```

```{r,echo=FALSE,eval=FALSE,message=FALSE,warning=FALSE}
# save
path <- paste0("../../results/UMAP/unannotated/",treatment,"_",tolower(tissue),
               "_UMAP_unannotated_split_treatment")
u1
saveToPDF(paste0(path, ".pdf"), width = 4, height = 4)
saveToPNG(paste0(path, ".png"), width = 4, height = 4, unit = "in", res = 300)
dev.off()

# save
path <- paste0("../../results/UMAP/unannotated/",treatment,"_",tolower(tissue),
               "_UMAP_unannotated_split_sample")
u2
saveToPDF(paste0(path, ".pdf"), width = 4, height = 4)
saveToPNG(paste0(path, ".png"), width = 4, height = 4, unit = "in", res = 300)
dev.off()

# save
path <- paste0("../../results/UMAP/unannotated/",treatment,"_",tolower(tissue),
               "_UMAP_unannotated_split_phase")
u3
saveToPDF(paste0(path, ".pdf"), width = 4, height = 4)
saveToPNG(paste0(path, ".png"), width = 4, height = 4, unit = "in", res = 300)
dev.off()

remove(u1,u2,u3)
```

## Revisit QC metrics
```{r umap_QC}
# nCount
f1 <- FeaturePlot(PMI.unannotated, 
            features = "nCount_RNA",
            pt.size = 0.4, 
            order = TRUE,
            min.cutoff = 'q10',
            label = TRUE)
f1

# nFeature
f2 <- FeaturePlot(PMI.unannotated, 
            features = "nFeature_RNA",
            pt.size = 0.4, 
            order = TRUE,
            min.cutoff = 'q10',
            label = TRUE)
f2

# percent.mt
f3 <- FeaturePlot(PMI.unannotated, 
            features = "percent.mt",
            pt.size = 0.4, 
            order = TRUE,
            min.cutoff = 'q10',
            label = TRUE)
f3

# cell.complexity
f4 <- FeaturePlot(PMI.unannotated, 
            features = "cell.complexity",
            pt.size = 0.4, 
            order = TRUE,
            min.cutoff = 'q10',
            label = TRUE)
f4

f5 <- FeaturePlot(PMI.unannotated, 
            features = "S.Score",
            pt.size = 0.4, 
            order = TRUE,
            min.cutoff = 'q10',
            label = TRUE)
f5

f6 <- FeaturePlot(PMI.unannotated, 
            features = "G2M.Score",
            pt.size = 0.4, 
            order = TRUE,
            min.cutoff = 'q10',
            label = TRUE)
f6
```

```{r,echo=FALSE,eval=FALSE,message=FALSE,warning=FALSE}
# save
path <- paste0("../../results/featurePlots/",treatment,"_",tolower(tissue),
               "_UMAP_unannotated_nCount")
f1
saveToPDF(paste0(path, ".pdf"), width = 4, height = 4)
saveToPNG(paste0(path, ".png"), width = 4, height = 4, unit = "in", res = 300)
dev.off()

# save
path <- paste0("../../results/featurePlots/",treatment,"_",tolower(tissue),
               "_UMAP_unannotated_nFeature")
f2
saveToPDF(paste0(path, ".pdf"), width = 4, height = 4)
saveToPNG(paste0(path, ".png"), width = 4, height = 4, unit = "in", res = 300)
dev.off()

# save
path <- paste0("../../results/featurePlots/",treatment,"_",tolower(tissue),
               "_UMAP_unannotated_percentMT")
f3
saveToPDF(paste0(path, ".pdf"), width = 4, height = 4)
saveToPNG(paste0(path, ".png"), width = 4, height = 4, unit = "in", res = 300)
dev.off()

# save
path <- paste0("../../results/featurePlots/",treatment,"_",tolower(tissue),
               "_UMAP_unannotated_complexity")
f4
saveToPDF(paste0(path, ".pdf"), width = 4, height = 4)
saveToPNG(paste0(path, ".png"), width = 4, height = 4, unit = "in", res = 300)
dev.off()

remove(f1,f2,f3,f4)
```

## Bulk DEGs
```{r}
# cell.complexity
f5 <- FeaturePlot(PMI.unannotated, 
            features = c("Xist"),
            pt.size = 0.4, 
            order = TRUE,
            min.cutoff = 'q1',
            label = FALSE)
f5

f6 <- FeaturePlot(PMI.unannotated, 
            features = "Malat1",
            pt.size = 0.4, 
            order = TRUE,
            min.cutoff = 'q1',
            label = TRUE)
f6
```

## Percent cells per cluster
```{r percent_cells_per_cluster}
PMI.unannotated@meta.data$seurat_clusters <- 
  PMI.unannotated@meta.data$SCT_snn_res.0.1

# treatment
b1 <- PMI.unannotated@meta.data %>%
  group_by(seurat_clusters, treatment) %>%
  dplyr::count() %>%
  group_by(seurat_clusters) %>%
  dplyr::mutate(percent = 100*n/sum(n)) %>%
  ungroup() %>%
  ggplot(aes(x=seurat_clusters,y=percent, fill=treatment)) +
  geom_col() +
  scale_fill_manual(values = treatment_colors) +
  ggtitle("Percentage of treatment per cluster")
b1

# sample
b2 <- PMI.unannotated@meta.data %>%
  group_by(seurat_clusters, sample) %>%
  dplyr::count() %>%
  group_by(seurat_clusters) %>%
  dplyr::mutate(percent = 100*n/sum(n)) %>%
  ungroup() %>%
  ggplot(aes(x=seurat_clusters,y=percent, fill=sample)) +
  geom_col() +
  #scale_fill_manual(values = sample_colors) +
  ggtitle("Percentage of sample per cluster")
b2
```

```{r,echo=FALSE,eval=FALSE}
# save
path <- paste0("../../results/nCells/",treatment,"_",tolower(tissue),
               "_percent_cells_per_cluster_treatment")
b1
saveToPDF(paste0(path, ".pdf"), width = 4, height = 4)
saveToPNG(paste0(path, ".png"), width = 4, height = 4, unit = "in", res = 300)
dev.off()

# save
path <- paste0("../../results/nCells/",treatment,"_",tolower(tissue),
               "_percent_cells_per_cluster_sample")
b2
saveToPDF(paste0(path, ".pdf"), width = 4, height = 4)
saveToPNG(paste0(path, ".png"), width = 4, height = 4, unit = "in", res = 300)
dev.off()
```

## Number cells per cluster
```{r cells_per_cluster}
treatment_ncells <- FetchData(PMI.unannotated, 
                     vars = c("ident", "treatment")) %>%
  dplyr::count(ident, treatment) %>%
  tidyr::spread(ident, n)
write.table(treatment_ncells, 
            paste0("../../results/nCells/",
                   treatment, "_",tolower(tissue),
                   "_cells_per_cluster_treatment.txt"),
            quote = FALSE, sep = "\t")

sample_ncells <- FetchData(PMI.unannotated, 
                     vars = c("ident", "sample")) %>%
  dplyr::count(ident,sample) %>%
  tidyr::spread(ident, n)
write.table(sample_ncells, 
            paste0("../../results/nCells/",
                   treatment, "_",tolower(tissue),
                   "_cells_per_cluster_sample.txt"),
            quote = FALSE, sep = "\t")
```

