---
title: "PMI - mice replicate group 1"
author: "Cadiz, Todd, Olney"
date: "04/14/2022"
output:
  html_document:
    theme: cerulean
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: inline
---

# Set working directory
```{r working_directory}
knitr::opts_knit$set(root.dir = ".")
```

# Load libraries
```{r libraries, message=FALSE, warnings=FALSE}
# load libraries
library(cowplot)  # plot_grid()
library(DoubletFinder)  # paramSweep()
library(dplyr)  # ungroup()
library(ggrepel) # geom_text_repel()
library(ggplot2)  # ggplot()
library(grid)  # grid.arrange()
library(gridExtra)  # grid.arrange()
library(harmony) # RunHarmony()
library(reshape2) # melt()
library(Seurat)  # Read10X_h5()
library(Seurat.utils) # RenameGenesSeurat()
library(stringr) # str_match()
```

# Set variables and thresholds
```{r set_variables}
pathToRef <- "/research/labs/neurology/fryer/projects/references/mouse/refdata-gex-mm10-2020-A"
tissue <- "Brain"
replicate <- "1"
treatment <- "PMI"
# genotype
WT <- "WT"
PS19 <- "PS19"
control <- "0_hr"
treatment_3hr <- "3_hr"
treatment_6hr <- "6_hr"
treatment_color_WT <- "blue"
treatment_color <- "blue"
control_color <- "gray"
WT_treatment_color_3hr <- "#6699FF" # light blue grey
WT_treatment_color_6hr <- "#0033FF" # dark blue
PS19_treatment_color_3hr <- "#CC99FF" # pink/purple
PS19_treatment_color_6hr <- "#9933FF" # purple
treatment_color_PS19 <- "#9933FF"
sample_colors <- c(control_color,WT_treatment_color_3hr, control_color, PS19_treatment_color_3hr)
treatment_colors <- c(control_color, treatment_color_WT, treatment_color_PS19)
myContrasts <- c("6 hour - 0 hour")
tool <- "cellranger"
```

```{r set_thresholds}
nCount.min <- 250
nFeature.min <- 250
complexity.cutoff <- 0.85
mt.cutoff <- 25
hb.cutoff <- 5
```

# Save functions
These functions with help simultaneously save plots as a png and pdf.
```{r save_functions}
saveToPDF <- function(...) {
    d = dev.copy(pdf,...)
    dev.off(d)
}
saveToPNG <- function(...) {
    d = dev.copy(png,...)
    dev.off(d)
}
```

# Setup seurat object
Using CellBender filtered output.
```{r seurat_object}
prefix <- "../../cellranger/snRNA/"
suffix <- "/outs/filtered_feature_bc_matrix.h5"

if (tissue == "Brain" && file.exists("../../rObjects/nucleus_replicate_1_merged_h5.rds")) {
  PMI <- readRDS("../../rObjects/nucleus_replicate_1_merged_h5.rds")
} else if (tissue == "Brain") {
  # individual sample objects
  WT_0_hr <- CreateSeuratObject(Read10X_h5(paste0(prefix,"WT_0_hr",suffix)))
  WT_3_hr <- CreateSeuratObject(Read10X_h5(paste0(prefix,"WT_3_hr",suffix)))
  #WT_6_hr <- CreateSeuratObject(Read10X_h5(paste0(prefix,"WT_6_hr",suffix)))
  PS19_0_hr <- CreateSeuratObject(Read10X_h5(paste0(prefix,"PS19_0_hr",suffix)))
  PS19_3_hr <- CreateSeuratObject(Read10X_h5(paste0(prefix,"PS19_3_hr",suffix)))
 # PS19_6_hr <- CreateSeuratObject(Read10X_h5(paste0(prefix,"PS19_6_hr",suffix)))
 
  # merge objects
  PMI <- merge(x = WT_0_hr, 
                y = c(WT_3_hr, PS19_0_hr, PS19_3_hr), 
                add.cell.ids = c("WT_0_hr", "WT_3_hr", "PS19_0_hr", "PS19_3_hr"), 
                project = paste0("PMI replicate group ", replicate, " - Single Cell"))
  
  # cleanup and save
  remove(WT_0_hr, WT_3_hr, PS19_0_hr, PS19_3_hr)
  saveRDS(PMI, "../../rObjects/nucleus_replicate_1_merged_h5.rds")
  
} 

# preview
PMI
```

nCount_RNA = total number of transcripts (UMIs) in a single cell 
nFeature_RNA = number of unique genes (features)

```{r change_idents}
# create sample column
barcodes <- colnames(PMI)
pattern <- "(.+)_[ACGT]+-(\\d+)"
sample <- str_match(barcodes, pattern)[,2]
table(sample)
PMI$sample <- factor(sample, levels = c("WT_0_hr", "WT_3_hr", "PS19_0_hr", "PS19_3_hr"))
table(PMI$sample)  # check
Idents(PMI) <- PMI$sample
```

Add treatment column to metadata
```{r treatment_column}
# create treatment column
treat <- gsub("WT_0_hr", control, PMI$sample)
treat <- gsub("WT_3_hr", treatment_3hr, treat)
#treat <- gsub("WT_6_hr", treatment_6hr, treat)
treat <- gsub("PS19_0_hr", control, treat)
treat <- gsub("PS19_3_hr", treatment_3hr, treat)
#treat <- gsub("PS19_6_hr", treatment_6hr, treat)

PMI$treatment <- factor(treat, levels = c("0_hr","3_hr"))
table(PMI$treatment)
```
Add genotype column to metadata
```{r genotype_column}
# create treatment column
geno <- gsub("WT_0_hr", WT, PMI$sample)
geno <- gsub("WT_3_hr", WT, geno)
#geno <- gsub("WT_6_hr", WT, geno)
geno <- gsub("PS19_0_hr", PS19, geno)
geno <- gsub("PS19_3_hr", PS19, geno)
#geno <- gsub("PS19_6_hr", PS19, geno)

PMI$genotype <- factor(geno, levels = c("WT","PS19"))
table(PMI$genotype)
```


```{r QC_columns}
# cell.complexity
PMI$cell.complexity <- log10(PMI$nFeature_RNA) / log10(PMI$nCount_RNA)

# percent MT
grep("^mt-",rownames(PMI@assays$RNA@counts),value = TRUE)
mt.genes <- c("mt-Nd1", "mt-Nd2", "mt-Co1", "mt-Co2", "mt-Atp8", "mt-Atp6", "mt-Co3", "mt-Nd3",
              "mt-Nd4l", "mt-Nd4", "mt-Nd5", "mt-Nd6", "mt-Cytb")
PMI$percent.mt <- PercentageFeatureSet(PMI, pattern = "^mt-")

# ribosomal proteins begin with 'RPL' or 'RSL' for this annotation file
PMI$percent.ribo <- PercentageFeatureSet(PMI, pattern = "^R[sp]l")

# percent.hb - hemoglobin proteins begin with 'HB' or 'HBP' 
PMI$percent.hb <- PercentageFeatureSet(PMI, pattern = "^Hb[^(p)]")

# percent xist in each cell 
PMI$percent.Xist <- PercentageFeatureSet(PMI, features = "Xist")

# percent xist in each cell 
PMI$percent.Ttr <- PercentageFeatureSet(PMI, features = "Ttr")
```

# Pre-filtering QC
## Number of cells
```{r prefiltering_cells_per_sample}
# Visualize the number of cell counts per sample
data <- as.data.frame(table(PMI$sample))
colnames(data) <- c("sample","frequency")

ncells1 <- ggplot(data, aes(x = sample, y = frequency, fill = sample)) + 
  geom_col() +
  theme_classic() +
  geom_text(aes(label = frequency), 
            position=position_dodge(width=0.9), 
            vjust=-0.25) +
  scale_fill_manual(values = sample_colors) + 
  scale_y_continuous(breaks = seq(0,55000, by = 5000), limits = c(0,55000)) +
  ggtitle("Raw: cells per sample") +
  theme(legend.position =  "none") + 
  theme(axis.text.x = element_text(angle = 45, hjust=1))
ncells1
```

## Density plots
```{r prefiltering_density}
# set graphical parameter
par(mfrow = c(3,1))

# Visualize nCount_RNA
den1 <- ggplot(PMI@meta.data,
       aes(color = sample,
           x = nCount_RNA,
           fill = sample)) +
  geom_density(alpha = 0.2) +
  theme_classic() +
  scale_x_log10() +
  scale_color_manual(values = sample_colors) +
  scale_fill_manual(values = sample_colors) +
  xlab("nCount_RNA") +
  ylab("Density") +
  geom_vline(xintercept = nCount.min)

# Visualize percent.mt
den2 <- ggplot(PMI@meta.data,
       aes(color = sample,
           x = percent.mt,
           fill = sample)) +
  geom_density(alpha = 0.2) +
  theme_classic() +
  scale_x_continuous(n.breaks = 4) +
  geom_vline(xintercept = mt.cutoff) +
  scale_color_manual(values = sample_colors) +
  scale_fill_manual(values = sample_colors) +
  xlab("% Mitochondrial Genes") +
  ylab("Density")

# Visualize cell complexity
# Quality cells are usually above 0.85
den3 <- ggplot(PMI@meta.data,
       aes(color = sample,
           x = cell.complexity,
           fill = sample)) +
  geom_density(alpha = 0.2) +
  theme_classic() +
  scale_color_manual(values = sample_colors) +
  scale_fill_manual(values = sample_colors) +
  xlab("Cell Complexity (log10(nFeature/nCount))") +
  ylab("Density") +
  geom_vline(xintercept = complexity.cutoff)

# Arrange graphs in grid
plots1 <- list(den1,den2,den3)
layout1 <- rbind(c(1),c(2),c(3))
grid1 <- grid.arrange(grobs = plots1, layout_matrix = layout1)

# Visualize percent.Ttr
den4 <- ggplot(PMI@meta.data,
       aes(color = sample,
           x = log2(percent.Ttr),
           fill = sample)) +
  geom_density(alpha = 0.2) +
  theme_classic() +
  scale_x_continuous(n.breaks = 4) +
 # geom_vline(xintercept = mt.cutoff) +
  scale_color_manual(values = sample_colors) +
  scale_fill_manual(values = sample_colors) +
  xlab("% Ttr Gene") +
  ylab("Density")
den4

# Visualize percent.Xist
den5 <- ggplot(PMI@meta.data,
       aes(color = sample,
           x = log2(percent.Xist),
           fill = sample)) +
  geom_density(alpha = 0.2) +
  theme_classic() +
  scale_x_continuous(n.breaks = 4) +
 # geom_vline(xintercept = mt.cutoff) +
  scale_color_manual(values = sample_colors) +
  scale_fill_manual(values = sample_colors) +
  xlab("% Xist Gene") +
  ylab("Density")
den5
```

## Violin plots
```{r prefiltering_violins}
# nFeature, nCount, and cell.complexity violins
v1 <- VlnPlot(PMI,
              features = c("nFeature_RNA", "nCount_RNA","cell.complexity"),
              ncol = 3,
              group.by = 'sample',
              cols = sample_colors,
              pt.size = 0)
v1

#  percent violins
v2 <- VlnPlot(PMI,
              features = c("percent.mt","percent.ribo","percent.hb", "percent.Xist"),
              ncol = 4,
              group.by = 'sample',
              cols = sample_colors,
              pt.size = 0)
v2

#  percent violins
vXist <- VlnPlot(PMI,
              features = c("percent.Xist"),
              group.by = 'sample',
              cols = sample_colors,
              pt.size = 0)
vXist

#  percent violins
vTtr <- VlnPlot(PMI,
              features = c("percent.Ttr"),
              group.by = 'sample',
              cols = sample_colors,
              pt.size = 0)
vTtr
```
# Histogram
```{r}
# User params
goi <- "Ttr"
sample_oi <- "PS19_3_hr"
threshold <- 2
# Split seurat object by timepoint to perform SCT on all samples
PMI.split <- SplitObject(PMI, split.by = "sample")

# Subset data
log2.threshold <- log2(threshold + 0.01)
counts.df <- FetchData(PMI.split$PS19_3_hr, vars = goi)
colnames(counts.df) <- "counts"
log2.counts.df <- log2(counts.df + 0.01)

# Histogram
title <- paste0("PMI Brain Cell: ", goi, " - ", sample_oi, "\nnCount_RNA > ", threshold)
hist1 <- ggplot(counts.df, aes(x = counts)) + 
  geom_histogram(bins = 100, fill = "gray", color = "black") + 
  labs(title = title, x=NULL, y=NULL) +
  xlab(paste0(goi, " nCount_RNA")) + ylab("# of cells") + theme_bw() +
  geom_vline(xintercept = threshold, col = "blue") +
  annotate("rect", 
           xmin = -Inf,
           xmax = threshold, 
           ymin = 0, 
           ymax=Inf, 
           alpha=0.2, 
           fill="deepskyblue") +
  annotate("rect", 
           xmin = threshold,
           xmax = Inf, 
           ymin = 0, 
           ymax=Inf, 
           alpha=0.2, 
           fill="chocolate4")

# Histogram log transformed
hist2 <- ggplot(log2.counts.df, aes(x = counts)) + 
  geom_histogram(bins = 100, fill = "gray", color = "black") + 
  labs(title = title, x=NULL, y=NULL) +
  xlab(paste0("Log2(",goi, " nCount_RNA)")) + ylab("# of cells") + theme_bw() +
  geom_vline(xintercept = log2.threshold, col = "blue") +
  annotate("rect", 
           xmin = -Inf,
           xmax = log2.threshold, 
           ymin = 0, 
           ymax=Inf, 
           alpha=0.2, 
           fill="deepskyblue") +
  annotate("rect", 
           xmin = log2.threshold,
           xmax = Inf, 
           ymin = 0, 
           ymax=Inf, 
           alpha=0.2, 
           fill="chocolate4")

# plot
plots1 <- list(hist1,hist2)
layout1 <- rbind(c(1),c(2))
grid1 <- grid.arrange(grobs = plots1, layout_matrix = layout1)

# number removed
table(counts.df$counts > threshold)

```