---
title: "SoupX"
author: "Kimberly Olney, PhD"
date: "04/14/2024"
output:
  html_document:
    theme: cerulean
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: inline
---

# SoupX to remove ambinet RNA
The [vignette](https://rawcdn.githack.com/constantAmateur/SoupX/204b602418df12e9fdb4b68775a8b486c6504fe4/inst/doc/pbmcTutorial.html)\
Estimate the composition of this soup, i.e. what fraction of UMIs are derived from the soup in each droplet and produce a corrected count table with the soup based expression removed.\

The method to do this consists of three parts:\

1. Calculate the profile of the soup.\
2. Estimate the cell specific contamination fraction.\
3. Infer a corrected expression matrix.\

```{r setup, include=FALSE}
knitr::opts_chunk$set(root.dir = ".", echo = TRUE, warning=FALSE, message=FALSE)
```

```{r libraries}
library(SeuratObject)
library(Seurat)
library(dplyr)
library(patchwork)
library(knitr)
library(xlsx)
library(ggplot2)
library(gridExtra)
library(grid)
library(lattice)
library(dittoSeq)
library(SoupX)
library(data.table)
Sys.setenv(RSTUDIO_PANDOC="/usr/local/biotools/pandoc/3.1.2/bin")
color.panel <- dittoColors()
```

The contamination fraction is parametrised as rho in the code, with rho=0 meaning no contamination and rho=1 meaning 100% of UMIs in a droplet are soup.
```{r soupX}
source(here::here("/research/labs/neurology/fryer/m239830/PMI/scripts", "file_paths_and_colours.R"))
treatment <- "PMI"
sampleID <- c("P_69", "B_7", "P_65", "B_6", "P_52", "P_62", "P_67", "J_3", "J_4", "P_60", "P_71", "B_10", "B_8", "B_9", "P_68", "P_54", "P_53", "P_64", "J_2", "J_1")
cellranger_results <- "/research/labs/neurology/fryer/m239830/PMI/cellranger/"

# rhoEst table
# rhoEst_table <- data.frame(sampleID = character(), rhoEst = numeric())

for (id in sampleID) {
  # Load data 
  sc <- load10X(paste0(cellranger_results, id, "/outs"))
  # Estimate the soup
  #sc <- autoEstCont(sc)
  #rhoEst <- sc$fit$rhoEst
  # Append sampleID and rhoEst to the dataframe
  #rhoEst_table <- rbind(rhoEst_table, data.frame(sampleID = id, rhoEst = rhoEst))
  # or set the contamination fraction to 20% for all cells.
  sc <- setContaminationFraction(sc, 0.2) 
  out <- adjustCounts(sc) # Adjust the counts 
  # Now that we've corrected our data, we can see how that compares to our uncorrected data. 
  # By default the function plots the fraction of expression in each cell that has been deemed to be soup.
  gene_names <- c("Aqp4", "Rbfox3", "Mog", "Vcan", "Ttr")
  for (gene_name in gene_names) {
    plot <- plotChangeMap(sc, out, gene_name) + theme_bw() + ggtitle(gene_name)
    ggsave(filename = paste0("../results/soupX/", gene_name, "_", id, ".pdf"), plot, width = 5, height = 4.5)
  }
  DropletUtils:::write10xCounts(paste0(cellranger_results, id, "/SoupX_set_fraction_20percent/"), out)
}

# Print the table
#print(rhoEst_table)
```
```{r}
```
