---
title: "doublet removal"
author: "Kimberly Olney, PhD"
date: "05/13/2024"
output:
  html_document:
    theme: cerulean
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: inline
---

Read in processed filtered data.
Remove doublets and reprocess. Check resolutions and save object. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(root.dir = ".", echo=TRUE, warning=FALSE, message=FALSE)
library(DoubletFinder)
library(reshape2)
```

# Libraris, paths, colors
```{r echo=FALSE, message=FALSE}
source(here::here("/research/labs/neurology/fryer/m239830/PMI/scripts", "file_paths_and_colours.R"))
treatment <- "PMI"
# Remove sample
metadata <- subset(metadata, sampleID != "P_60"  & sampleID != "B_6")
color.panel <- dittoColors()
```

# Read in object
```{r read_object}
# read object
dataObject <- readRDS(file = paste0("../rObjects/", treatment, ".annotated.rds"))
#DefaultAssay(dataObject) <- "RNA"
#dataObject <- NormalizeData(dataObject)
#dataObject <- FindVariableFeatures(dataObject)
#dataObject <- ScaleData(dataObject)
#dataObject <- JoinLayers(dataObject)

# inspect
dataObject
```

# Unannotated 
### UMAP
```{r unannotated_umap}
ditto_umap <- dittoDimPlot(object = dataObject,
             var = "individual_clusters",
             reduction.use = "umap",
             do.label = TRUE,
             labels.highlight = TRUE)
ditto_umap
```

### Cluster tree
```{r cluster_tree}
dataObject <- BuildClusterTree(
  object = dataObject,
  dims = 1:30,
  reorder = FALSE,
  reorder.numeric = FALSE
)

tree <- dataObject@tools$BuildClusterTree
tree$tip.label <- paste0("Cluster ", tree$tip.label)
nClusters <- length(tree$tip.label)

tree_graph <- ggtree::ggtree(tree, aes(x, y)) +
  scale_y_reverse() +
  ggtree::geom_tree() +
  ggtree::theme_tree() +
  ggtree::geom_tiplab(offset = 1) +
  ggtree::geom_tippoint(color = color.panel[1:nClusters],
                        shape = 16,
                        size = 5) +
  coord_cartesian(clip = 'off') +
  theme(plot.margin = unit(c(0, 2.5, 0, 0), 'cm'))
tree_graph
```

```{r save_tree}
# save
path <- paste0("../results/UMAP/unannotated/", treatment,
               "_annotated_tree")
tree_graph
saveToPDF(paste0(path, ".pdf"), width = 5.5, height = 4)
```
## Doublet filtering
https://github.com/chris-mcginnis-ucsf/DoubletFinder\

Heterotypic - doublets derived from transcriptionally distinct cells. 
DoubletFinder works best on this type of doublet. \

Homotopic - Transcriptionally similar cell doublets. DoubletFinder does not work 
as great on this type of doublet.\

pANN - proportion of artificial nearest neighbors (pANN)\

BCMVN - mean-variance normalized bimodality coefficient of pANN distributions 
produced during pN -pK parameter sweeps. The BCMVN may be used to identify the 
pK parameter.\

Overview of steps:\
A. Prepare each sample\
B. pK Identification (no ground-truth) - defines the PC neighborhood size used 
   to compute pANN\
C. Homotypic Doublet Proportion Estimate - homotypic doublets may not be a 
   problem depending on the type of analysis you are performing. If you have 
   some doublets of the same type and their counts are normalized, they will 
   generally represent the profile of single cells of the same type.\
D. DoubletFinder\
E. Visualize where the doublets are located\

```{r split_object}
# split object by sample
dataObject.split <- SplitObject(dataObject, split.by = "sample") 
```

```{r doubletFinder, message=FALSE, warning=FALSE}
for (i in 1:length(dataObject.split)) {
  # normalize and find PCs
  print(i)
  PMI_sample <- NormalizeData(dataObject.split[[i]])
  sampleID <- levels(droplevels(PMI_sample@meta.data$sample))
  PMI_sample <- FindVariableFeatures(PMI_sample, selection.method = "vst", nfeatures = 2000)
  PMI_sample <- ScaleData(PMI_sample)
  PMI_sample <- RunPCA(PMI_sample)
  
  # get significant PCs
  stdv <- PMI_sample[["pca"]]@stdev
  sum.stdv <- sum(PMI_sample[["pca"]]@stdev)
  percent.stdv <- (stdv / sum.stdv) * 100
  cumulative <- cumsum(percent.stdv)
  co1 <- which(cumulative > 90 & percent.stdv < 5)[1]
  co2 <- sort(which((percent.stdv[1:length(percent.stdv) - 1] - 
                       percent.stdv[2:length(percent.stdv)]) > 0.1), 
              decreasing = T)[1] + 1
  min.pc <- min(co1, co2)
  min.pc
  
  # run umap
  PMI_sample <- RunUMAP(PMI_sample, dims = 1:min.pc, reduction = "pca")
  
  # cluster
  PMI_sample <- FindNeighbors(object = PMI_sample, dims = 1:min.pc)                           
  PMI_sample <- FindClusters(object = PMI_sample, resolution = 0.2)
  
  # Assign identity of clusters
  Idents(object = PMI_sample) <- "seurat_clusters"

  # number of cells in each cluster
  n_cells <- FetchData(PMI_sample, vars = c("ident")) %>% dplyr::count(ident) %>%tidyr::spread(ident, n)
  
  ## pK Identification (no ground-truth) 
  sweep.res.list <- paramSweep(PMI_sample, PCs = 1:min.pc, sct = FALSE)
  sweep.stats <- summarizeSweep(sweep.res.list, GT = FALSE)
  bcmvn <- find.pK(sweep.stats)
  
  # Optimal pK for any scRNA-seq data can be manually discerned as maxima in BCmvn distributions
  bcmvn_max <- bcmvn[which.max(bcmvn$BCmetric),]
  pK_value <- bcmvn_max$pK
  pK_value <- as.numeric(levels(pK_value))[pK_value]
  
  # Homotypic Doublet Proportion Estimate 
  annotations <- PMI_sample@meta.data$individual_clusters
  homotypic.prop <- modelHomotypic(annotations) 
  nExp_poi <- round(pK_value*nrow(PMI_sample@meta.data))
  nExp_poi.adj <- round(nExp_poi*(1-homotypic.prop))
  # Run DoubletFinder with varying classification  
  PMI_sample <- doubletFinder(PMI_sample, PCs = 1:min.pc, 
                pN = 0.25, pK = pK_value, nExp = nExp_poi.adj, 
                reuse.pANN = FALSE, sct = FALSE)
  
  # set DF class for calling doublets
  DF_class <- PMI_sample@meta.data[, grep("DF.classifications",colnames(PMI_sample@meta.data)),]
  DF_class[which(DF_class == "Doublet")] <- "Doublet"
  table(DF_class)
  
  # table showing the number of doublets and singlets
  write.table(table(DF_class), paste0("../results/doublet_finder/",treatment, "_doubletFinder_table_",sampleID), sep = "\t", 
              row.names = FALSE, quote = FALSE)
  PMI_sample@meta.data[,"CellTypes_DF"] <- DF_class
  
  # plot
  d2 <- DimPlot(PMI_sample, group.by="CellTypes_DF", reduction="umap",
          order=c("Coll.Duct.TC","Doublet"), 
          cols=c("#66C2A5","black"))
  path <- paste0("../results/doublet_finder/",treatment,
               "_doubletFinder_UMAP_",sampleID)
  pdf(paste0(path, ".pdf"), width = 5,height = 4)
  print(d2)
  dev.off()
  
  # plot
  f1 <- FeaturePlot(PMI_sample, 
            reduction = "umap", 
            features = c("nFeature_RNA", "nCount_RNA", 
                         "cell.complexity", "percent.mt"),
            pt.size = 0.4, 
            order = TRUE,
            label = TRUE)
  path <- paste0("../results/doublet_finder/",treatment,
               "_doubletFinder_FeaturePlot_",sampleID)
  pdf(paste0(path, ".pdf"), width = 10, height = 7)
  print(f1)
  dev.off()
  
  #only keep singlets
  PMI_sample_singlets <- subset(PMI_sample, subset = CellTypes_DF == "Singlet")
  

  # number of cells in each cluster per and post removing doublets
  n_cells_singlets <- FetchData(PMI_sample_singlets, vars = c("ident")) %>% dplyr::count(ident) %>% tidyr::spread(ident, n)
  n_cells_singlets
  ncells_per_cluster <- rbind(n_cells, n_cells_singlets)
  row.names(ncells_per_cluster) <- c("Doublets and singlets", "Singlets only")
  ncells_per_cluster
  difference <- diff(as.matrix(ncells_per_cluster))
  difference <- as.data.frame(difference)
  row.names(difference) <- c("difference")
  cbind(difference, ncells_per_cluster)
  write.table(ncells_per_cluster, paste0(
    "../results/doublet_finder/",treatment,
    "_doubletFinder_table_ncells_per_cluster_",sampleID, ".txt"), sep = "\t", 
    row.names = FALSE, quote = FALSE)
  # plot the number of cells in each cluster per and post doubletFinder
  ncell_matrix <- as.matrix(ncells_per_cluster)
  ncells_melt <- melt(ncell_matrix)
  colnames(ncells_melt) <- c("doublet type","cluster","number of cells")
  ncell_max <- ncells_melt[which.max(ncells_melt$`number of cells`),]
  ncell_max_value <- ncell_max$`number of cells`
  cellmax <- ncell_max_value + 800 # so that the figure doesn't cut off the text
  b1 <- ggplot(ncells_melt, aes(x = factor(cluster), y = `number of cells`,
                          fill = `doublet type`)) + 
    geom_bar(stat="identity", colour="black", width=1, position = position_dodge(width=0.8)) +
    geom_text(aes(label = `number of cells`), 
              position=position_dodge(width=0.9), vjust=-0.25, angle = 45, hjust=-.01) + 
    theme_classic() + scale_fill_manual(values = c("gray", "#66C2A5")) +
    ggtitle("Number of cells per cluster") +  xlab("cluster") +
    theme(axis.text.x = element_text(angle = 45, hjust=1)) +
    scale_y_continuous(limits = c(0,cellmax))
  path <- paste0("../results/doublet_finder/",treatment,
               "_doubletFinder_barplot_ncells_per_cluster_",sampleID)
  pdf(paste0(path, ".pdf"), width = 7,height = 5)
  print(b1)
  dev.off()
  f2 <- FeaturePlot(PMI_sample_singlets, 
            reduction = "umap", 
            features = c("nFeature_RNA", "nCount_RNA", 
                         "cell.complexity", "percent.mt"),
            pt.size = 0.4, 
            order = TRUE,
            label = TRUE)
  path <- paste0("../results/doublet_finder/",treatment,
               "_doubletFinder_FeaturePlot_singlets_",sampleID)
  pdf(paste0(path, ".pdf"), width = 10,height = 7)
  print(f2)
  dev.off()
  
  # put the PMI together again
  dataObject.split[[i]] <- PMI_sample_singlets
}
```

```{r}
# converge PMI.split
dataObject.singlets <- merge(x = dataObject.split[[1]],
                       y = c(dataObject.split[[2]], dataObject.split[[3]], dataObject.split[[4]], dataObject.split[[5]], dataObject.split[[6]],
                       dataObject.split[[7]], dataObject.split[[8]], dataObject.split[[9]], dataObject.split[[10]], dataObject.split[[11]],
                       dataObject.split[[12]], dataObject.split[[13]], dataObject.split[[14]], dataObject.split[[15]], dataObject.split[[16]],
                       dataObject.split[[17]], dataObject.split[[18]]),
                       project = paste0("PMI"))

# print how many cells removed
print(paste0(dim(dataObject)[2] - dim(dataObject.singlets)[2]," nuclie removed"))

# how many removed if we had an upper nCount and nFeature
dataObject.upper <- subset(dataObject.singlets,
                     subset = (nCount_RNA < 20000) & (nFeature_RNA < 8000))
print(paste0(dim(dataObject.singlets)[2] - dim(dataObject.upper)[2],
             " nuclei would have been removed if upper bound applied"))

```

# Save singlets
```{r}
saveRDS(dataObject.upper, paste0("../rObjects/",treatment,".singlets.rds"))
dataObject.upper <- readRDS(paste0("../rObjects/",treatment,".singlets.rds"))
```

```{r cleanup}
# cleanup
remove(dataObject.singlets, dataObject.split, dataObject)
remove(n_cells,n_cells_singlets,ncell_matrix,ncell_max,ncells_per_cluster,ncells_melt)
remove(sweep.res.list, sweep.stats,bcmvn,bcmvn_max,difference)
remove(d1,d2,d3,f1,f2)
remove(counts,counts.filtered, nonzero)

dataObject <- dataObject.upper
```


# Singlets
### DotPlot
```{r dot_individual}
markers.to.plot <-
  c(
"Clu", 
"Gfap", 
"Aqp4", 
"Gja1", 
"Cldn5", 
"Adgrf5", 
"Flt1", 
"Col1a1", 
"Col1a2", 
"Dcn", 
"Hexb", 
"C1qa", 
"C1qc", 
"C1qb", 
"Tmem119", 
"Itgam", 
"Tyrobp", 
"P2ry12", 
"Aif1", 
"Rbfox3", 
"Snap25", 
"Syt1", 
"Gad1", 
"Gad2", 
"Plp1", 
"Mbp", 
"Mog", 
"Olig1", 
"Pdgfra", 
"Vcan", 
"Tnr", 
"Acta2", 
"Rgs5", 
"Vtn", 
"Myl5"
  )

dot_ind <- DotPlot(dataObject.upper,
                   features = markers.to.plot, 
                  # split.by = "strain", 
                   cluster.idents = TRUE,
                   dot.scale = 8) + RotatedAxis()
dot_ind
```



# Reprocess post-doublet removal 
```{r reprocess}
# save singlets as the new dataObject
dataObject <- dataObject.upper

# transform
dataObject <- SCTransform(dataObject, verbose = FALSE)

# run PCA on the merged object
dataObject <- RunPCA(object = dataObject)
Idents(dataObject) <- "sample"

# Determine the K-nearest neighbor graph
dataObject <- FindNeighbors(object = dataObject, 
                                 assay = "SCT", 
                                 reduction = "pca",
                                 dims = 1:15)
# Run UMAP
dataObject <- RunUMAP(dataObject,
                           dims = 1:15,
                           reduction = "pca",
                           n.components = 3) 

# Determine the clusters for various resolutions
dataObject <- FindClusters(object = dataObject,
                                 algorithm = 1, # 1= Louvain
                                 resolution = seq(0.1,.6,by=0.1))
```

### Explore resolutions
```{r umap_noDoublets}
Idents(dataObject) <- dataObject$SCT_snn_res.0.6
dataObject$seurat_clusters <- dataObject$SCT_snn_res.0.6
ditto_umap <- dittoDimPlot(object = dataObject,
             var = "seurat_clusters",
             reduction.use = "umap",
             do.label = TRUE,
             labels.highlight = TRUE)
ditto_umap

# 0.4
umap0.2 <- DimPlot(dataObject,
        group.by = "SCT_snn_res.0.2",
        label = TRUE)
umap0.2

# 0.4
umap0.4 <- DimPlot(dataObject,
        group.by = "SCT_snn_res.0.4",
        label = TRUE)
umap0.4
```

```{r save_umap_noDoublets}
# save
path <- paste0("../results/UMAP/unannotated/",treatment,
               "_UMAP_unannotated_doublets_removed")
ditto_umap
saveToPDF(paste0(path, ".pdf"), width = 7, height = 6.6)
```

## Nuclei count per cluster
```{r nuclei_per_cluster_noDoublets}
Idents(dataObject) <- dataObject$seurat_clusters
sample_ncells <- FetchData(dataObject, 
                     vars = c("ident", "sample")) %>%
  dplyr::count(ident,sample) %>%
  tidyr::spread(ident, n)
write.table(sample_ncells, 
            paste0("../results/nuclei_count/",
                   treatment, 
                   "_nuclei_per_cluster_doublets_removed.txt"),
            quote = FALSE, sep = "\t")
sample_ncells
```

### DotPlot
```{r dot_individual}
markers.to.plot <-
  c(
"Clu", 
"Gfap", 
"Aqp4", 
"Gja1", 
"Cldn5", 
"Adgrf5", 
"Flt1", 
"Col1a1", 
"Col1a2", 
"Dcn", 
"Hexb", 
"C1qa", 
"C1qc", 
"C1qb", 
"Tmem119", 
"Itgam", 
"Tyrobp", 
"P2ry12", 
"Aif1", 
"Rbfox3", 
"Snap25", 
"Syt1", 
"Gad1", 
"Gad2", 
"Plp1", 
"Mbp", 
"Mog", 
"Olig1", 
"Pdgfra", 
"Vcan", 
"Tnr", 
"Acta2", 
"Rgs5", 
"Vtn", 
"Myl5"
  )

dot_ind <- DotPlot(dataObject,
                   features = markers.to.plot, 
                  # split.by = "strain", 
                   cluster.idents = TRUE,
                   dot.scale = 8) + RotatedAxis()
dot_ind
```

```{r save_dot_individual, echo=FALSE}
pdf(
  paste0(
    "../results/dot_plot/",
    treatment,
    "_clusters_DotPlot_no_integration_doublets_removed.pdf"
  ),
  width = 14,
  height = 10
)
dot_ind
dev.off()
```
# Save
```{r save_object,echo=FALSE,eval=TRUE}
saveRDS(dataObject, paste0("../rObjects/",treatment,"_unannotated_doublets_removed.rds"))
```

```{r}
```

