---
title: "Wilcox differential expression"
author: "Kimberly Olney, PhD"
date: "07/08/2024"
output:
  html_document:
    theme: cerulean
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: inline
---

Differential expression was completed for the following comparisons:
1.	PS19 vs WT within each cell type. All samples.\
2.	3 hours versus 0 hours within each cell type. All samples.\ 
3.	3 hours versus 0 hours within each cell type. PS19 only.\   
4.	3 hours versus 0 hours within each cell type. WT only.\ 
5.	3 hours versus 0 hours within each cell type. XX female only.\ 
6.	3 hours versus 0 hours within each cell type. XY male only.\ 
7.	PS19 vs WT within each cell type. 0 hours only.\
8.	PS19 vs WT within each cell type. 3 hours only.\

```{r setup, include=FALSE}
knitr::opts_chunk$set(root.dir = ".", echo=TRUE, warning=FALSE, message=FALSE)
```

# Libraris, paths, colors
https://satijalab.org/seurat/articles/essential_commands.html 
```{r echo=FALSE, message=FALSE}
source(here::here("/research/labs/neurology/fryer/m239830/PMI/scripts", "file_paths_and_colours.R"))
treatment <- "PMI"
color.panel <- dittoColors()
metadata <- subset(metadata, sampleID != "P_60" & sampleID != "B_6")
metadata$sampleID <- factor(metadata$sampleID, levels = c(metadata$sampleID))
samples <- metadata$sampleID 
```

# Read in object
```{r read_object}
# read object
dataObject <- readRDS(file = paste0("../rObjects/", treatment, ".dataObject.clean.rds"))
# inspect
dataObject
Layers(dataObject)
# List of cell types to analyze
cell_types <- c(unique(dataObject$individual_clusters))

# only protein coding genes
protein_coding_genes <- subset(gene_info, type == "protein_coding")
counts <- GetAssayData(object = dataObject, layer = "counts")
keep <- rownames(counts) %in% protein_coding_genes$gene_name # false when mt.gene
counts.filtered <- counts[keep,]

# overwrite dataObject.filtered
dataObject <- CreateSeuratObject(counts.filtered, 
                                    meta.data = dataObject@meta.data)

# print features removed
print(paste0(dim(counts)[1] - dim(counts.filtered)[1], " features removed"))

DefaultAssay(dataObject) <- "RNA"
dataObject <- NormalizeData(dataObject)
```

# Wilcox 
## 1) PS19 vs WT
## 2) 3 hours vs Fresh
```{r Wilcox_DE_all_samples}
# Directory to save the results
output_dir <- "../results/DEGs/Wilcox/"

# Loop through each cell type to perform differential expression analysis for each condition
for (cell_type in cell_types) {
  # Strain comparison
  dataObject$celltype.strain <- paste(dataObject$individual_clusters, dataObject$strain, sep = "_")
  Idents(dataObject) <- "celltype.strain"
  df <- FindMarkers(dataObject, ident.1 = paste(cell_type, "PS19", sep = "_"), ident.2 = paste(cell_type, "WT", sep = "_"), 
                                  verbose = FALSE)
  df_sig <- subset(df, df$p_val_adj < 0.05)
  df$percent_difference <- abs(df$pct.1 - df$pct.2) # percent difference 
  colnames(df)[c(3,4)] <- c("percent_PS19","percent_WT") # define the percent columns 
  df$gene <- rownames(df) # set row names as column gene 
  DEGs <- df[,c(7,1,2,5,3,4,6)] # reorder 
  strain_output_path <- file.path(output_dir, paste0(cell_type, "_PS19_vs_WT_comparison.txt"))
  write.table(DEGs, file = strain_output_path, sep = "\t", quote = FALSE, row.names = FALSE)

  # Time comparison
  dataObject$celltype.time <- paste(dataObject$individual_clusters, dataObject$time, sep = "_")
  Idents(dataObject) <- "celltype.time"
  df <- FindMarkers(dataObject, ident.1 = paste(cell_type, "3", sep = "_"), ident.2 = paste(cell_type, "0", sep = "_"), 
                                verbose = FALSE)
  # Save the time comparison results to a text file
  df$percent_difference <- abs(df$pct.1 - df$pct.2)
  colnames(df)[c(3,4)] <- c("percent_3","percent_0")
  df$gene <- rownames(df)
  DEGs <- df[,c(7,1,2,5,3,4,6)]
  time_output_path <- file.path(output_dir, paste0(cell_type, "_3_vs_0_hours_comparison.txt"))
  write.table(DEGs, file = time_output_path, sep = "\t", quote = FALSE, row.names = FALSE)
  rm(df, DEGs)
}
```

# DE 0 hours only: PS19 vs WT
```{r DE_0hours_only}
Idents(dataObject) <- dataObject$time
dataObject.0hours <- subset(dataObject, cells = WhichCells(dataObject, idents = setdiff(unique(dataObject$time), c(3))))
Idents(dataObject.0hours) <- dataObject.0hours$individual_clusters # Re-define the idents as the cell type
output_dir <- "../results/DEGs/Wilcox/"
for (cell_type in cell_types) {
  dataObject.0hours$celltype.strain <- paste(dataObject.0hours$individual_clusters, dataObject.0hours$strain, sep = "_")
  Idents(dataObject.0hours) <- "celltype.strain"
  df <- FindMarkers(dataObject.0hours, ident.1 = paste(cell_type, "PS19", sep = "_"), ident.2 = paste(cell_type, "WT", sep = "_"), 
                                    verbose = FALSE)
  df$percent_difference <- abs(df$pct.1 - df$pct.2) # percent difference 
  colnames(df)[c(3,4)] <- c("percent_PS19","percent_WT") # define the percent columns 
  df$gene <- rownames(df) # set row names as column gene 
  DEGs <- df[,c(7,1,2,5,3,4,6)] # reorder 
  strain_output_path <- file.path(output_dir, paste0(cell_type, "_PS19_vs_WT_comparison_0hours_only.txt"))
  write.table(DEGs, file = strain_output_path, sep = "\t", quote = FALSE, row.names = FALSE)
  rm(df, DEGs)
}
rm(dataObject.0hours)
```

# DE 3 hours only: PS19 vs WT
```{r DE_3hours_only}
Idents(dataObject) <- dataObject$time
dataObject.3hours <- subset(dataObject, cells = WhichCells(dataObject, idents = setdiff(unique(dataObject$time), c(0))))
Idents(dataObject.3hours) <- dataObject.3hours$individual_clusters # Re-define idents as cell types 
output_dir <- "../results/DEGs/Wilcox/"
for (cell_type in cell_types) {
  dataObject.3hours$celltype.strain <- paste(dataObject.3hours$individual_clusters, dataObject.3hours$strain, sep = "_")
  Idents(dataObject.3hours) <- "celltype.strain"
  df <- FindMarkers(dataObject.3hours, ident.1 = paste(cell_type, "PS19", sep = "_"), ident.2 = paste(cell_type, "WT", sep = "_"), 
                                   verbose = FALSE)
  df$percent_difference <- abs(df$pct.1 - df$pct.2) # percent difference 
  colnames(df)[c(3,4)] <- c("percent_PS19","percent_WT") # define the percent columns 
  df$gene <- rownames(df) # set row names as column gene 
  DEGs <- df[,c(7,1,2,5,3,4,6)] # reorder 
  strain_output_path <- file.path(output_dir, paste0(cell_type, "_PS19_vs_WT_comparison_3hours_only.txt"))
  write.table(DEGs, file = strain_output_path, sep = "\t", quote = FALSE, row.names = FALSE)
  rm(df, DEGs)
}
rm(dataObject.3hours)
```

# DE PS19 only: 3 hours vs Fresh
```{r DE_PS19_only}
Idents(dataObject) <- dataObject$strain
dataObject.PS19 <- subset(dataObject, cells = WhichCells(dataObject, idents = setdiff(unique(dataObject$strain), c("WT"))))
Idents(dataObject.PS19) <- dataObject.PS19$individual_clusters 
output_dir <- "../results/DEGs/Wilcox/"

for (cell_type in cell_types) {
  dataObject.PS19$celltype.time <- paste(dataObject.PS19$individual_clusters, dataObject.PS19$time, sep = "_")
  Idents(dataObject.PS19) <- "celltype.time"
  df <- FindMarkers(dataObject.PS19, ident.1 = paste(cell_type, "3", sep = "_"), ident.2 = paste(cell_type, "0", sep = "_"), 
                                   verbose = FALSE)
  df$percent_difference <- abs(df$pct.1 - df$pct.2) # percent difference 
  colnames(df)[c(3,4)] <- c("percent_3","percent_0") # define the percent columns 
  df$gene <- rownames(df) # set row names as column gene 
  DEGs <- df[,c(7,1,2,5,3,4,6)] # reorder 
  strain_output_path <- file.path(output_dir, paste0(cell_type, "_3_vs_0_hours_comparison_PS19_only.txt"))
  write.table(DEGs, file = strain_output_path, sep = "\t", quote = FALSE, row.names = FALSE)
  rm(df, DEGs)
}
```

# DE WT only: 3 hours vs Fresh
```{r DE_WT_only}
Idents(dataObject) <- dataObject$strain
dataObject.WT <- subset(dataObject, cells = WhichCells(dataObject, idents = setdiff(unique(dataObject$strain), c("PS19"))))
Idents(dataObject.WT) <- dataObject.WT$individual_clusters
output_dir <- "../results/DEGs/Wilcox/"
for (cell_type in cell_types) {
  # Strain comparison
  dataObject.WT$celltype.time <- paste(dataObject.WT$individual_clusters, dataObject.WT$time, sep = "_")
  Idents(dataObject.WT) <- "celltype.time"
  df <- FindMarkers(dataObject.WT, ident.1 = paste(cell_type, "3", sep = "_"), ident.2 = paste(cell_type, "0", sep = "_"), 
                                   verbose = FALSE)
  df$percent_difference <- abs(df$pct.1 - df$pct.2) # percent difference 
  colnames(df)[c(3,4)] <- c("percent_3","percent_0") # define the percent columns 
  df$gene <- rownames(df) # set row names as column gene 
  DEGs <- df[,c(7,1,2,5,3,4,6)] # reorder 
  output_path <- file.path(output_dir, paste0(cell_type, "_3_vs_0_hours_comparison_WT_only.txt"))
  write.table(DEGs, file = output_path, sep = "\t", quote = FALSE, row.names = FALSE)
  rm(df, DEGs, output_path)
}
rm(dataObject.WT, output_dir)
```

# DE XX female only: 3 hours vs Fresh
```{r DE_XX_female_only}
Idents(dataObject) <- dataObject$sex
dataObject.female <- subset(dataObject, cells = WhichCells(dataObject, idents = setdiff(unique(dataObject$sex), c("Male"))))
Idents(dataObject.female) <- dataObject.female$individual_clusters

output_dir <- "../results/DEGs/Wilcox/"
for (cell_type in cell_types) {
  dataObject.female$celltype.time <- paste(dataObject.female$individual_clusters, dataObject.female$time, sep = "_")
  Idents(dataObject.female) <- "celltype.time"
  df <- FindMarkers(dataObject.female, ident.1 = paste(cell_type, "3", sep = "_"), ident.2 = paste(cell_type, "0", sep = "_"), 
                                  verbose = FALSE)
  df$percent_difference <- abs(df$pct.1 - df$pct.2) # percent difference 
  colnames(df)[c(3,4)] <- c("percent_3","percent_0") # define the percent columns 
  df$gene <- rownames(df) # set row names as column gene 
  DEGs <- df[,c(7,1,2,5,3,4,6)] # reorder 
  output_path <- file.path(output_dir, paste0(cell_type, "_3_vs_0_hours_comparison_XX_female_only.txt"))
  write.table(DEGs, file = output_path, sep = "\t", quote = FALSE, row.names = FALSE)
  rm(df, DEGs, output_path)
}
rm(dataObject.female, output_dir)
```

# DE XY male only: 3 hours vs Fresh
```{r DE_XY_male_only}
Idents(dataObject) <- dataObject$sex
dataObject.male <- subset(dataObject, cells = WhichCells(dataObject, idents = setdiff(unique(dataObject$sex), c("Female"))))
Idents(dataObject.male) <- dataObject.male$individual_clusters
output_dir <- "../results/DEGs/Wilcox/"
for (cell_type in cell_types) {
  dataObject.male$celltype.time <- paste(dataObject.male$individual_clusters, dataObject.male$time, sep = "_")
  Idents(dataObject.male) <- "celltype.time"
  df <- FindMarkers(dataObject.male, ident.1 = paste(cell_type, "3", sep = "_"), ident.2 = paste(cell_type, "0", sep = "_"), 
                                   verbose = FALSE)
  df$percent_difference <- abs(df$pct.1 - df$pct.2) # percent difference 
  colnames(df)[c(3,4)] <- c("percent_3","percent_0") # define the percent columns 
  df$gene <- rownames(df) # set row names as column gene 
  DEGs <- df[,c(7,1,2,5,3,4,6)] # reorder 
  output_path <- file.path(output_dir, paste0(cell_type, "_3_vs_0_hours_comparison_XY_male_only.txt"))
  write.table(DEGs, file = output_path, sep = "\t", quote = FALSE, row.names = FALSE)
  rm(df, DEGs, output_path)
}
rm(dataObject.male, output_dir)
```



