---
title: "DESeq2 differential expression"
author: "Kimberly Olney, PhD"
date: "05/28/2024"
output:
  html_document:
    theme: cerulean
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: inline
---

Differential expression was completed for the following comparisons:
1.	PS19 vs WT within each cell type. All samples.\
2.	3 hours versus 0 hours within each cell type. All samples.\ 
3.	3 hours versus 0 hours within each cell type. PS19 only.\   
4.	3 hours versus 0 hours within each cell type. WT only.\ 
5.	3 hours versus 0 hours within each cell type. XX female only.\ 
6.	3 hours versus 0 hours within each cell type. XY male only.\ 
7.	PS19 vs WT within each cell type. 0 hours only.\
8.	PS19 vs WT within each cell type. 3 hours only.\

```{r setup, include=FALSE}
knitr::opts_chunk$set(root.dir = ".", echo=TRUE, warning=FALSE, message=FALSE)
```

# Libraris, paths, colors
https://satijalab.org/seurat/articles/essential_commands.html 
```{r echo=FALSE, message=FALSE}
source(here::here("/research/labs/neurology/fryer/m239830/PMI/scripts", "file_paths_and_colours.R"))
treatment <- "PMI"
color.panel <- dittoColors()
metadata <- subset(metadata, sampleID != "P_60" & sampleID != "B_6")
metadata$sampleID <- factor(metadata$sampleID, levels = c(metadata$sampleID))
samples <- metadata$sampleID 
```

# Read in object
```{r read_object}
# read object
dataObject <- readRDS(file = paste0("../rObjects/", treatment, ".dataObject.clean.rds"))
# inspect
dataObject
Layers(dataObject)
# List of cell types to analyze
cell_types <- c(unique(dataObject$individual_clusters))
```
```{r protein_coding}
protein_coding_genes <- subset(gene_info, type == "protein_coding")
```

# Check 
```{r }
# Group
table(dataObject$group) 
```

# DESeq all samples 
```{r DESeq_all_samples}
info <- (dataObject@meta.data)

output_dir <- "../results/DEGs/FindMarkers_DESeq2_pseudobulk"
# List of cell types to analyze
cell_types <- c(unique(dataObject$individual_clusters))
# Pseudo-bulk the counts based on sample, strain, sex, time, and cell type
dataObject.pseudo <- AggregateExpression(
    dataObject, 
    assays = "RNA", # DESeq works with raw counts
    features = protein_coding_genes$gene_name,
    return.seurat = TRUE, 
    group.by = c("sample", "strain", "sex", "time", "individual_clusters")
)

# Loop through each cell type to perform differential expression analysis for each condition
for (cell_type in cell_types) {
    # Strain comparison with covariates
    dataObject.pseudo$celltype.strain <- paste(dataObject.pseudo$individual_clusters, dataObject.pseudo$strain, sep = "_")
    Idents(dataObject.pseudo) <- "celltype.strain"
    DEGs <- FindMarkers(object = dataObject.pseudo, 
                         ident.1 = paste0(cell_type, "_PS19"), 
                         ident.2 = paste0(cell_type, "_WT"),
                         test.use = "DESeq2")
  DEGs$percent_difference <- abs(DEGs$pct.1 - DEGs$pct.2) # percent difference 
  colnames(DEGs)[c(3,4)] <- c("percent_PS19","percent_WT") # define the percent columns 
  DEGs$gene <- rownames(DEGs) # set row names as column gene 
  DEGs <- DEGs[,c(7,1,2,5,3,4,6)] # reorder 
    strain_output_path <- file.path(output_dir, paste0(cell_type, "_PS19_vs_WT_comparison_pseudobulk.txt"))
    write.table(DEGs, file = strain_output_path, sep = "\t", quote = FALSE, row.names = FALSE)
    rm(dds, res, DEGs) # clean up 
    
    # Time comparison with covariates
    dataObject.pseudo$celltype.time <- paste(dataObject.pseudo$individual_clusters, dataObject.pseudo$time, sep = "_")
    Idents(dataObject.pseudo) <- "celltype.time"
    DEGs <- FindMarkers(object = dataObject.pseudo, 
                         ident.1 = paste0(cell_type, "_3"), 
                         ident.2 = paste0(cell_type, "_0"),
                         test.use = "DESeq2")
      DEGs$percent_difference <- abs(DEGs$pct.1 - DEGs$pct.2) # percent difference 
  colnames(DEGs)[c(3,4)] <- c("percent_3","percent_0") # define the percent columns 
  DEGs$gene <- rownames(DEGs) # set row names as column gene 
  DEGs <- DEGs[,c(7,1,2,5,3,4,6)] # reorder 
    time_output_path <- file.path(output_dir, paste0(cell_type, "_3_vs_0_hours_comparison_pseudobulk.txt"))
    write.table(DEGs, file = time_output_path, sep = "\t", quote = FALSE, row.names = FALSE)
    rm(DEGs) # clean up 
}
rm(dataObject.pseudo)
```


# DESeq within a group 
#### PS19 vs WT
```{r}
output_dir <- "../results/DEGs/FindMarkers_DESeq2_pseudobulk"

# Function to construct the output filename
construct_filename <- function(cell_type, comparison, subgroup) {
    paste0(cell_type, "_", comparison, "_comparison_", subgroup, ".txt")
}

# List of conditions to subset data
conditions <- list(
    list(subset_expr = "time == '0'", group_by = "strain", comparison = "PS19_vs_WT", subgroup = "0_hours_only", design_formula = ~ sex + strain, DES_group_by = c("sample", "strain", "sex", "individual_clusters")),
    list(subset_expr = "time == '3'", group_by = "strain", comparison = "PS19_vs_WT", subgroup = "3_hours_only", design_formula = ~ sex + strain, DES_group_by = c("sample", "strain", "sex", "individual_clusters")),
    list(subset_expr = "sex == 'Female'", group_by = "strain", comparison = "PS19_vs_WT", subgroup = "XX_female_only", design_formula = ~ time + strain, DES_group_by = c("sample", "strain", "time", "individual_clusters")),
    list(subset_expr = "sex == 'Male'", group_by = "strain", comparison = "PS19_vs_WT", subgroup = "XY_male_only", design_formula = ~ time + strain, DES_group_by = c("sample", "strain", "time", "individual_clusters"))
)

# Loop through each cell type and condition
for (cell_type in cell_types) {
    for (condition in conditions) {
        cat("Processing condition:", condition$subset_expr, "\n")
        
        # Subset dataObject based on the condition using logical indexing
        subset_indices <- with(dataObject@meta.data, eval(parse(text = condition$subset_expr)))
        dataObject_subset <- subset(dataObject, cells = rownames(dataObject@meta.data)[subset_indices])
        
        cat("Number of cells after subsetting:", nrow(dataObject_subset@meta.data), "\n")
        
        if (nrow(dataObject_subset@meta.data) > 0) {
            # Pseudo-bulk the counts for the subsetted data
            dataObject.pseudo <- AggregateExpression(
                dataObject_subset, 
                features = protein_coding_genes$gene_name,
                assays = "RNA", 
                return.seurat = TRUE, 
                group.by = condition$DES_group_by
            )

            # Define the identifier for the current cell type
            dataObject.pseudo$celltype.strain <- paste(dataObject.pseudo$individual_clusters, dataObject.pseudo$strain, sep = "_")
            Idents(dataObject.pseudo) <- "celltype.strain"
            DEGs <- FindMarkers(object = dataObject.pseudo, 
                         ident.1 = paste0(cell_type, "_PS19"), 
                         ident.2 = paste0(cell_type, "_WT"),
                         test.use = "DESeq2")
        DEGs$percent_difference <- abs(DEGs$pct.1 - DEGs$pct.2) # percent difference 
        colnames(DEGs)[c(3,4)] <- c("percent_PS19","percent_WT") # define the percent columns 
        DEGs$gene <- rownames(DEGs) # set row names as column gene 
        DEGs <- DEGs[,c(7,1,2,5,3,4,6)] # reorder 
                
                # Construct the output filename
                output_filename <- construct_filename(cell_type, condition$comparison, condition$subgroup)
                output_path <- file.path(output_dir, output_filename)
                write.table(DEGs, file = output_path, sep = "\t", quote = FALSE, row.names = FALSE)
                rm(dds, res, DEGs) # clean up 
    }}}
rm(dataObject_subset, dataObject.pseudo)

```

#### 3 vs 0 hours
```{r}
output_dir <- "../results/DEGs/FindMarkers_DESeq2_pseudobulk"

# Function to construct the output filename
construct_filename <- function(cell_type, comparison, subgroup) {
    paste0(cell_type, "_", comparison, "_comparison_", subgroup, ".txt")
}


# List of conditions to subset data
conditions <- list(
    list(subset_expr = "strain == 'PS19'", group_by = "time", comparison = "3_vs_0_hours", subgroup = "WT_only", design_formula = ~ sex + time, DES_group_by = c("sample", "time", "sex", "individual_clusters")),
    list(subset_expr = "strain == 'WT'", group_by = "time", comparison = "3_vs_0_hours", subgroup = "PS19_only", design_formula = ~ sex + time, DES_group_by = c("sample", "time", "sex", "individual_clusters")))

# Loop through each cell type and condition
for (cell_type in cell_types) {
    for (condition in conditions) {
        cat("Processing condition:", condition$subset_expr, "\n")
        
        # Subset dataObject based on the condition using logical indexing
        subset_indices <- with(dataObject@meta.data, eval(parse(text = condition$subset_expr)))
        dataObject_subset <- subset(dataObject, cells = rownames(dataObject@meta.data)[subset_indices])
        
        cat("Number of cells after subsetting:", nrow(dataObject_subset@meta.data), "\n")
        
        if (nrow(dataObject_subset@meta.data) > 0) {
            # Pseudo-bulk the counts for the subsetted data
            dataObject.pseudo <- AggregateExpression(
                dataObject_subset, 
                features = protein_coding_genes$gene_name,
                assays = "RNA", 
                return.seurat = TRUE, 
                group.by = condition$DES_group_by
            )

            # Define the identifier for the current cell type
            dataObject.pseudo$celltype.time <- paste(dataObject.pseudo$individual_clusters, dataObject.pseudo$time, sep = "_")
            Idents(dataObject.pseudo) <- "celltype.time"
            DEGs <- FindMarkers(object = dataObject.pseudo, 
                         ident.1 = paste0(cell_type, "_3"), 
                         ident.2 = paste0(cell_type, "_0"),
                         test.use = "DESeq2")
        DEGs$percent_difference <- abs(DEGs$pct.1 - DEGs$pct.2) # percent difference 
        colnames(DEGs)[c(3,4)] <- c("percent_3","percent_0") # define the percent columns 
        DEGs$gene <- rownames(DEGs) # set row names as column gene 
        DEGs <- DEGs[,c(7,1,2,5,3,4,6)] # reorder 
                
                # Construct the output filename
                output_filename <- construct_filename(cell_type, condition$comparison, condition$subgroup)
                output_path <- file.path(output_dir, output_filename)
                write.table(DEGs, file = output_path, sep = "\t", quote = FALSE, row.names = FALSE)
                rm(dds, res, DEGs) # clean up 
    }}}
rm(dataObject_subset, dataObject.pseudo)
```

# PCA post pseudo-bulk 
```{r}
# Step 1: Pseudo-bulk the counts based on sample, strain, sex, time, and cell type
dataObject.pseudo <- AggregateExpression(
    dataObject, 
    assays = "RNA", 
    features = protein_coding_genes$gene_name,
    return.seurat = TRUE, 
    group.by = c("sample", "strain", "sex", "time", "individual_clusters")
)

# Step 2: Normalize the data
dataObject.pseudo <- NormalizeData(dataObject.pseudo, normalization.method = "LogNormalize", scale.factor = 10000)

# Step 3: Find variable features
dataObject.pseudo <- FindVariableFeatures(dataObject.pseudo, selection.method = "vst", nfeatures = 2000)

# Step 4: Scale the data
dataObject.pseudo <- ScaleData(dataObject.pseudo, features = rownames(dataObject.pseudo))

# Step 5: Run PCA
dataObject.pseudo <- RunPCA(dataObject.pseudo, features = VariableFeatures(object = dataObject.pseudo))

# Step 6: Visualize PCA
DimPlot(dataObject.pseudo, reduction = "pca", label = TRUE) + NoLegend()

pdf(paste0("../results/pca/PMI_DESeq2_psuedobulk_cell_type.pdf"), height = 5, width = 6.5)
DimPlot(dataObject.pseudo, reduction = "pca", group.by = "individual_clusters")
dev.off()

# Step 7: Extract PC scores and metadata
pc_scores <- as.data.frame(Embeddings(dataObject.pseudo, reduction = "pca")[, 1:10]) # Extract PC1 to PC10
# Step 7: Extract metadata
metadata <- as.data.frame(dataObject.pseudo@meta.data)

# Step 8: Combine PC scores and metadata
combined_data <- cbind(pc_scores, metadata)

library(variancePartition)

# Define the formula for CCA
form_PCA <- ~ time +
  sex +
  strain +
  individual_clusters +
  PC_1 +
  PC_2 +
  PC_3 +
  PC_4 +
  PC_5 +
  PC_6 +
  PC_7 +
  PC_8 +
  PC_9 +
  PC_10 

# Perform Canonical Correlation Analysis
C <- canCorPairs(form_PCA, combined_data)

# Define a function to compute p-values for correlation
cor.mtest <- function(C, ...) {
    C <- as.matrix(C)
    n <- ncol(C)
    p.mat <- matrix(NA, n, n)
    diag(p.mat) <- 0
    for (i in 1:(n - 1)) {
        for (j in (i + 1):n) {
            tmp <- cor.test(C[, i], C[, j], ...)
            p.mat[i, j] <- p.mat[j, i] <- tmp$p.value
        }
    }
    colnames(p.mat) <- rownames(p.mat) <- colnames(C)
    p.mat
}

# Compute p-values for correlation
p.mat <- cor.mtest(C)

# Define colors for the heatmap
col <- colorRampPalette(c("#4477AA", "#77AADD", "#FFFFFF", "#EE9988", "#BB4444"))
library(corrplot)
# Plot correlation matrix using corrplot
corrplot(C, method = "color", col = col(200),  
         type = "upper", order = "hclust", 
         addCoef.col = "black", 
         tl.col = "black", tl.srt = 45, 
         diag = FALSE, col.lim = c(0, 1))

# PCA 
DimPlot(dataObject.pseudo, reduction = "pca", group.by = "individual_clusters")

```



## Other 
```{r}
library(dittoSeq)
dittoDimPlot(dataObject, "sex")
library(SummarizedExperiment)

pseudo_bulk_counts <- GetAssayData(dataObject.pseudo, slot = "counts")

# Alternate accessor function with the same result
pseudo_bulk_meta <- dataObject.pseudo@meta.data

pseudo_bulk_meta$sex

bulkSE <- SummarizedExperiment(
    assays = list(counts = pseudo_bulk_counts),
    colData = data.frame(pseudo_bulk_meta)
)

# Create the heatmap
dittoHeatmap(dataObject,
             genes_to_visualize,
             annot.by = c("time", "sex"), 
             order.by = "time",  # This orders the columns by the "time" annotation
             scaled.to.max = FALSE,
             show_colnames = FALSE,
             show_rownames = TRUE)

help(dittoHeatmap)
dittoBarPlot(bulkSE, "sex", group.by = "conditions")
```

```{r}
library(Seurat)
library(pheatmap)

# Get unique cell types
cell_types <- unique(dataObject$individual_clusters)

# Initialize a list to store the top genes for each cell type
top_genes_list <- list()

# Iterate over cell types
for (cell_type in cell_types) {
  # Subsetting by cell type
  celltype_data <- subset(dataObject, ident = cell_type)
  
  # Calculate average expression across all cells in this cell type
  avg_expression <- AverageExpression(celltype_data)
  
  # Order genes by average expression
  top_genes <- names(sort(avg_expression, decreasing = TRUE))[1:10]
  
  # Add top genes to the list
  top_genes_list[[cell_type]] <- top_genes
}

# Get the union of top genes across all cell types
all_top_genes <- unique(unlist(top_genes_list))

# Create a matrix to store the expression data for top genes
heatmap_data <- matrix(NA, nrow = length(all_top_genes), ncol = ncol(dataObject), dimnames = list(all_top_genes, colnames(dataObject)))

# Fill in the matrix with expression data
for (cell_type in cell_types) {
  # Subsetting by cell type
  celltype_data <- subset(seurat_object, ident = cell_type)
  
  # Get the top genes for this cell type
  top_genes <- top_genes_list[[cell_type]]
  
  # Fill in the expression data for top genes in this cell type
  heatmap_data[top_genes, colnames(seurat_object)[IdentifySeuratObject(celltype_data)]] <- celltype_data@assays$RNA@data[top_genes, ]
}

# Generate the heatmap
pheatmap(heatmap_data,
         cluster_rows = FALSE,
         cluster_cols = FALSE,
         show_rownames = TRUE,
         show_colnames = TRUE,
         main = "Top 10 most highly expressed genes in each cell type")
```