---
title: "Metascape inputs"
author: "Kimberly Olney, PhD"
date: "07/16/2024"
output:
  html_document:
    theme: cerulean
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: inline
---

Differential expression was completed for the following comparisons:
1.	PS19 vs WT within each cell type. All samples.\
2.	3 hours versus 0 hours within each cell type. All samples.\ 
3.	3 hours versus 0 hours within each cell type. PS19 only.\   
4.	3 hours versus 0 hours within each cell type. WT only.\ 
5.	3 hours versus 0 hours within each cell type. XX female only.\ 
6.	3 hours versus 0 hours within each cell type. XY male only.\ 
7.	PS19 vs WT within each cell type. 0 hours only.\
8.	PS19 vs WT within each cell type. 3 hours only.\

```{r setup, include=FALSE}
knitr::opts_chunk$set(root.dir = ".", echo=TRUE, warning=FALSE, message=FALSE)
```

# Libraris, paths, colors
https://satijalab.org/seurat/articles/essential_commands.html 
```{r echo=FALSE, message=FALSE}
source(here::here("/research/labs/neurology/fryer/m239830/PMI/scripts", "file_paths_and_colours.R"))
treatment <- "PMI"
color.panel <- dittoColors()
metadata <- subset(metadata, sampleID != "P_60" & sampleID != "B_6")
metadata$sampleID <- factor(metadata$sampleID, levels = c(metadata$sampleID))
samples <- metadata$sampleID 
cell_types <- c("neuron", "interneuron", "astrocyte", "oligodendrocyte", "polydendrocyte", "endothelial", "mural", "fibroblast", "microglia")
```

## Read excel table of significant DEGs
Make a table for each cell type and each comparison. This is what will be read into Metascape. 
```{r time_comparison}
#------------------- All samples 
for (i in cell_types) {
  filepath <- paste0("../results/DEGs/DESeq2_pseudobulk_exp_filter/", i, "_3_vs_0_hours_comparison_pseudobulk.txt")
  dataset <- read.delim(filepath, header = TRUE, sep = "\t", stringsAsFactors = FALSE)
  up_filtered_dataset <- subset(dataset, padj < 0.1 & log2FoldChange > 0)
  write.table(up_filtered_dataset$gene, paste0("../results/metascape_input/", i, "_3_vs_0_hours_up_regulated.txt"), row.names = FALSE, quote = FALSE, col.names = FALSE)
  down_filtered_dataset <- subset(dataset, padj < 0.1 & log2FoldChange < 0)
  write.table(down_filtered_dataset$gene, paste0("../results/metascape_input/", i, "_3_vs_0_hours_down_regulated.txt"), row.names = FALSE, quote = FALSE, col.names = FALSE)
}

#------------------- PS19 only
for (i in cell_types) {
  filepath <- paste0("../results/DEGs/DESeq2_pseudobulk_exp_filter/", i, "_3_vs_0_hours_comparison_PS19_only.txt")
  dataset <- read.delim(filepath, header = TRUE, sep = "\t", stringsAsFactors = FALSE)
  up_filtered_dataset <- subset(dataset, padj < 0.1 & log2FoldChange > 0)
  write.table(up_filtered_dataset$gene, paste0("../results/metascape_input/", i, "_3_vs_0_hours_up_regulated_PS19_only.txt"), row.names = FALSE, quote = FALSE, col.names = FALSE)
  down_filtered_dataset <- subset(dataset, padj < 0.1 & log2FoldChange < 0)
  write.table(down_filtered_dataset$gene, paste0("../results/metascape_input/", i, "_3_vs_0_hours_down_regulated_PS19_only.txt"), row.names = FALSE, quote = FALSE, col.names = FALSE)
}
#------------------- WT only
for (i in cell_types) {
  filepath <- paste0("../results/DEGs/DESeq2_pseudobulk_exp_filter/", i, "_3_vs_0_hours_comparison_WT_only.txt")
  dataset <- read.delim(filepath, header = TRUE, sep = "\t", stringsAsFactors = FALSE)
  up_filtered_dataset <- subset(dataset, padj < 0.1 & log2FoldChange > 0)
  write.table(up_filtered_dataset$gene, paste0("../results/metascape_input/", i, "_3_vs_0_hours_up_regulated_WT_only.txt"), row.names = FALSE, quote = FALSE, col.names = FALSE)
  down_filtered_dataset <- subset(dataset, padj < 0.1 & log2FoldChange < 0)
  write.table(down_filtered_dataset$gene, paste0("../results/metascape_input/", i, "_3_vs_0_hours_down_regulated_WT_only.txt"), row.names = FALSE, quote = FALSE, col.names = FALSE)
}
```

#### PS19 vs WT
```{r PS19_vs_WT}
#------------------- All samples
for (i in cell_types) {
  filepath <- paste0("../results/DEGs/DESeq2_pseudobulk_exp_filter/", i, "_PS19_vs_WT_comparison_pseudobulk.txt")
  dataset <- read.delim(filepath, header = TRUE, sep = "\t", stringsAsFactors = FALSE)
  up_filtered_dataset <- subset(dataset, padj < 0.1 & log2FoldChange > 0)
  write.table(up_filtered_dataset$gene, paste0("../results/metascape_input/", i, "_PS19_vs_WT_up_regulated.txt"), row.names = FALSE, quote = FALSE, col.names = FALSE)
  down_filtered_dataset <- subset(dataset, padj < 0.1 & log2FoldChange < 0)
  write.table(down_filtered_dataset$gene, paste0("../results/metascape_input/", i, "_PS19_vs_WT_down_regulated.txt"), row.names = FALSE, quote = FALSE, col.names = FALSE)
}

#------------------- 3 hours only 
for (i in cell_types) {
  filepath <- paste0("../results/DEGs/DESeq2_pseudobulk_exp_filter/", i, "_PS19_vs_WT_comparison_3_hours_only.txt")
  dataset <- read.delim(filepath, header = TRUE, sep = "\t", stringsAsFactors = FALSE)
  up_filtered_dataset <- subset(dataset, padj < 0.1 & log2FoldChange > 0)
  write.table(up_filtered_dataset$gene, paste0("../results/metascape_input/", i, "_PS19_vs_WT_up_regulated_3_hours_only.txt"), row.names = FALSE, quote = FALSE, col.names = FALSE)
  down_filtered_dataset <- subset(dataset, padj < 0.1 & log2FoldChange < 0)
  write.table(down_filtered_dataset$gene, paste0("../results/metascape_input/", i, "_PS19_vs_WT_down_regulated_3_hours_only.txt"), row.names = FALSE, quote = FALSE, col.names = FALSE)
}

#------------------- 0 hours only 
for (i in cell_types) {
  filepath <- paste0("../results/DEGs/DESeq2_pseudobulk_exp_filter/", i, "_PS19_vs_WT_comparison_0_hours_only.txt")
  dataset <- read.delim(filepath, header = TRUE, sep = "\t", stringsAsFactors = FALSE)
  up_filtered_dataset <- subset(dataset, padj < 0.1 & log2FoldChange > 0)
  write.table(up_filtered_dataset$gene, paste0("../results/metascape_input/", i, "_PS19_vs_WT_up_regulated_0_hours_only.txt"), row.names = FALSE, quote = FALSE, col.names = FALSE)
  down_filtered_dataset <- subset(dataset, padj < 0.1 & log2FoldChange < 0)
  write.table(down_filtered_dataset$gene, paste0("../results/metascape_input/", i, "_PS19_vs_WT_down_regulated_0_hours_only.txt"), row.names = FALSE, quote = FALSE, col.names = FALSE)
}

```

