---
title: "Correlation in the log2FC comparisons"
author: "Kimberly Olney, PhD"
date: "05/29/2024"
output:
  html_document:
    theme: cerulean
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: inline
---

Differential expression was completed for the following comparisons:
1.	PS19 vs WT within each cell type. All samples.\
2.	PS19 vs WT within each cell type. 0 hours only.\
3.	PS19 vs WT within each cell type. 3 hours only.\

```{r setup, include=FALSE}
knitr::opts_chunk$set(root.dir = ".", echo=TRUE, warning=FALSE, message=FALSE)
library(ComplexHeatmap)
library(ggpubr)
```

# Libraris, paths, colors
https://satijalab.org/seurat/articles/essential_commands.html 
```{r echo=FALSE, message=FALSE}
source(here::here("/research/labs/neurology/fryer/m239830/PMI/scripts", "file_paths_and_colours.R"))
treatment <- "PMI"
color.panel <- dittoColors()
metadata <- subset(metadata, sampleID != "P_60" & sampleID != "B_6")
metadata$sampleID <- factor(metadata$sampleID, levels = c(metadata$sampleID))
samples <- metadata$sampleID 
```

# Read in object
```{r read_object}
# read object
dataObject <- readRDS(file = paste0("../rObjects/", treatment, ".dataObject.clean.rds"))
# inspect
dataObject
Layers(dataObject)
# List of cell types to analyze
cell_types <- c(unique(dataObject$individual_clusters))
```
# Sex check
```{r exp_matrix}
# Pseudo-bulk the counts based on sample, strain, sex, time, and cell type
dataObject.pseudo <- AggregateExpression(
    dataObject, 
    assays = "RNA", # DESeq works with raw counts
    return.seurat = TRUE, 
    group.by = c("sample", "strain", "sex", "time", "individual_clusters")
)
# Normalize the data
dataObject.pseudo <- NormalizeData(dataObject.pseudo, normalization.method = "LogNormalize", scale.factor = 10000)
help(NormalizeData)
# Find variable features
dataObject.pseudo <- FindVariableFeatures(dataObject.pseudo, selection.method = "vst", nfeatures = 2000)

# Scale the data
dataObject.pseudo <- ScaleData(dataObject.pseudo)

# Perform PCA
dataObject.pseudo <- RunPCA(dataObject.pseudo, features = VariableFeatures(object = dataObject.pseudo))

# Visualize the expression of a specific gene
FeaturePlot(dataObject.pseudo, features = "Rbfox3")

# Visualize the distribution of expression levels
pdf(paste0("../results/sex_check/violin_Xist.pdf"), width = 8.5, height = 4)
VlnPlot(dataObject.pseudo, features = "Xist", group.by = "sample")
dev.off()

pdf(paste0("../results/sex_check/violin_Uty.pdf"), width = 8.5, height = 4)
VlnPlot(dataObject.pseudo, features = "Uty", group.by = "sample")
dev.off()


pseudo_bulk_counts <- GetAssayData(dataObject.pseudo, layer = "counts")
pseudo_bulk_counts <- NormalizeData(pseudo_bulk_counts)
df_norm <- as.matrix(pseudo_bulk_counts)

Uty <- subset(df_norm, rownames(df_norm) == "Uty")
Xist <- subset(df_norm, rownames(df_norm) == "Xist")

library(reshape2)
genes <- rbind(Uty, Xist)
genes_melt <- melt(genes)

# Extract sample ID
genes_melt$sampleID <- sapply(strsplit(as.character(genes_melt$Var2), '_'), '[', 1)
genes_melt$reported_sex <- sapply(strsplit(as.character(genes_melt$Var2), '_'), '[', 3)

# Reshape data to wide format for plotting
data_wide <- reshape(genes_melt, timevar = "Var1", idvar = "Var2", direction = "wide")

# Specify the sample ID to be highlighted
highlight_sample <- "P-68"

# Add a column for coloring
data_wide$color <- ifelse(data_wide$sampleID.Uty == highlight_sample, highlight_sample, "Other")

# Plot the data
ggplot(data_wide, aes(x = value.Xist, y = value.Uty, color = color)) +
  geom_point(size = 3) +
  scale_color_manual(values = c("gray", "red")) + # Highlight the specific sample in red and others in black
  labs(x = "Xist Expression", y = "Uty Expression") +
  theme_minimal() +
  theme(legend.title = element_blank())

```

# Expression heatmap for each cell type, all expressed genes 
```{r heatmap_celltype, eval=FALSE}
# Define colors for each level of qualitative variables
col = list(
  strain = c("PS19" = "orange", "WT" = "yellow"),
  sex = c("Male" = "green", "Female" = "purple"),
  timepoint = c("3" = "red", "0" = "blue")
)

# Loop over each cell type
for (cell_type in cell_types) {
  # Filter the data for the current cell type
  df_filtered <- df_norm[, grep(paste0("_", cell_type, "$"), colnames(df_norm))]
  
   # Filter genes with counts above 1 in at least 3 samples
  gene_filter <- rowSums(df_filtered > 1) >= 3
  df_filtered <- df_filtered[gene_filter, ]
  
  # Extract annotations from column names
  annotations <- data.frame(
    strain = sapply(strsplit(colnames(df_filtered), "_"), function(x) x[2]), 
    sex = sapply(strsplit(colnames(df_filtered), "_"), function(x) x[3]),
    timepoint = sapply(strsplit(colnames(df_filtered), "_"), function(x) x[4])
  )
  
  # Create the heatmap annotation
  ha <- HeatmapAnnotation(
    df = annotations,
    col = col
  )
  
  # Create the heatmap
  heatmap_plot <- Heatmap(df_filtered, name = "Expression",
                          top_annotation = ha,
                          show_row_names = FALSE)
  
  # Save the heatmap as a PDF
  pdf(paste0("../results/heatmap/", cell_type, "_expression.pdf"),
      width = 8.5,
      height = 10)
  draw(heatmap_plot)
  dev.off()
}
```

# 3 vs 0 hours DEGs heatmap
```{r heatmap_3_vs_0}
# Define colors for each level of qualitative variables
col = list(
  strain = c("PS19" = "orange", "WT" = "yellow"),
  sex = c("Male" = "green", "Female" = "purple"),
  timepoint = c("3" = "red", "0" = "blue")
)

# Loop over each cell type
for (cell_type in cell_types) {
  # Filter the data for the current cell type
  df_filtered <- df_norm[, grep(paste0("_", cell_type, "$"), colnames(df_norm))]

  DEGs <- read.delim(paste0("../results/DEGs/DESeq2_pseudobulk/", cell_type,
                            "_3_vs_0_hours_comparison_pseudo_bulk_with_covariates.txt"))
  gene_filter <- subset(DEGs, (padj < 0.05 & log2FoldChange > 0.25) | ((padj < 0.05 & log2FoldChange < -0.25)))
  df_filtered <- df_filtered[gene_filter$gene, ]
  
  # Extract annotations from column names
  annotations <- data.frame(
    strain = sapply(strsplit(colnames(df_filtered), "_"), function(x) x[2]), 
    sex = sapply(strsplit(colnames(df_filtered), "_"), function(x) x[3]),
    timepoint = sapply(strsplit(colnames(df_filtered), "_"), function(x) x[4])
  )
  
  # Create the heatmap annotation
  ha <- HeatmapAnnotation(
    df = annotations,
    col = col
  )
  
  # Create the heatmap
  heatmap_plot <- Heatmap(df_filtered, name = "Expression",
                          top_annotation = ha,
                          show_row_names = TRUE, # show gene names
                          column_names_rot = 75)  # Rotate column names

  
  # Save the heatmap as a PDF
  pdf(paste0("../results/heatmap/", cell_type, "_expression_3_vs_0_hours_DEGs_q0.05_log2FC0.25.pdf"),
      width = 14,
      height = 10)
  draw(heatmap_plot)
  dev.off()
}
```

# PS19 vs WT  DEGs heatmap
```{r heatmap_PS19_vs_WT}
# Define colors for each level of qualitative variables
col = list(
  strain = c("PS19" = "orange", "WT" = "yellow"),
  sex = c("Male" = "green", "Female" = "purple"),
  timepoint = c("3" = "red", "0" = "blue")
)

# Loop over each cell type
for (cell_type in cell_types) {
  # Filter the data for the current cell type
  df_filtered <- df_norm[, grep(paste0("_", cell_type, "$"), colnames(df_norm))]

  DEGs <- read.delim(paste0("../results/DEGs/DESeq2_pseudobulk/", cell_type,
                            "_PS19_vs_WT_comparison_pseudo_bulk_with_covariates.txt"))
  gene_filter <- subset(DEGs, (padj < 0.05 & log2FoldChange > 0.25) | ((padj < 0.05 & log2FoldChange < -0.25)))
  df_filtered <- df_filtered[gene_filter$gene, ]
  
  # Extract annotations from column names
  annotations <- data.frame(
    strain = sapply(strsplit(colnames(df_filtered), "_"), function(x) x[2]), 
    sex = sapply(strsplit(colnames(df_filtered), "_"), function(x) x[3]),
    timepoint = sapply(strsplit(colnames(df_filtered), "_"), function(x) x[4])
  )
  
  # Create the heatmap annotation
  ha <- HeatmapAnnotation(
    df = annotations,
    col = col
  )
  
  # Create the heatmap
  heatmap_plot <- Heatmap(df_filtered, name = "Expression",
                          top_annotation = ha,
                          show_row_names = TRUE, # show gene names
                          column_names_rot = 75)  # Rotate column names

  
  # Save the heatmap as a PDF
  pdf(paste0("../results/heatmap/", cell_type, "_expression_PS19_vs_WT_DEGs_q0.05_log2FC0.25.pdf"),
      width = 14,
      height = 10)
  draw(heatmap_plot)
  dev.off()
}
```

  
```{r}
#----------
# Extract annotations from column names
annotations <- data.frame(
  strain = sapply(strsplit(colnames(df), "_"), function(x) x[2]),
  sex = sapply(strsplit(colnames(df), "_"), function(x) x[3]),
  timepoint = sapply(strsplit(colnames(df), "_"), function(x) x[4])
)

# Create the heatmap annotation
ha <- HeatmapAnnotation(
  df = annotations,
  col = col
)

# Plot the heatmap
heatmap_plot <- Heatmap(df, name = "Expression",
        top_annotation = ha,
        show_row_names = FALSE)

pdf(paste0("../results/heatmap/neuron_expression.pdf"
  ),
  width = 8.5,
  height = 8
)
heatmap_plot
dev.off()
```


# Correlation matrix
```{r}
# Load necessary libraries
library(dplyr)

# Define the list of cell types and time points
time_points <- c("0hours", "3hours")

# Initialize an empty list to store data frames
deg_list <- list()

# Loop through each cell type and time point to read the data
for (cell_type in cell_types) {
  for (time_point in time_points) {
    # Construct the file name
    file_name <- paste0("../results/DEGs/MAST/", cell_type, "_PS19_vs_WT_comparison_", time_point, "_only.txt")
    
    # Read the data
    deg_data <- read.table(file_name, header = TRUE, sep = "\t")
    
    # Add columns for cell type and time point
    deg_data <- deg_data %>%
      mutate(cell_type = cell_type, time_point = time_point)
    
    # Store the data frame in the list
    deg_list[[paste0(cell_type, "_", time_point)]] <- deg_data
  }
}

# Combine all data frames into a single data frame
deg_combined <- bind_rows(deg_list)

# Filter the data for 0 hours
deg_0hours <- deg_combined %>%
  filter(time_point == "0hours") %>%
  select(gene, avg_log2FC, cell_type)

# Filter the data for 0 hours
deg_3hours <- deg_combined %>%
  filter(time_point == "3hours") %>%
  select(gene, avg_log2FC, cell_type)

DEG_combined <- merge(deg_0hours, deg_3hours, by = c("gene", "cell_type"), suffixes = c("_0h", "_3h"))
head(DEG_combined)

# Compute correlation matrix by cell_type
correlation_results <- DEG_combined %>%
  group_by(cell_type) %>%
  summarise(correlation = cor(avg_log2FC_0h, avg_log2FC_3h))
```


# Plot the data
```{r}
ggplot(DEG_combined, aes(x = avg_log2FC_0h, y = avg_log2FC_3h)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  facet_wrap(~ cell_type, ncol = 3) +
  stat_cor(method = "pearson", label.x = -4, label.y = 5) +
  labs(
    title = "Correlation of 0 hours and 3 hours by cell type",
    x = "0 hours log2FC",
    y = "3 hours log2FC"
  ) +
  theme_minimal()
```

